---
title: "Compare State Abundance"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

## Introduction
The purpose of this workflow is to compare state abundance across models.
State abundance is calculated in 1.4_Chromatin_States_and_Expression.Rmd.

We use the emissions of the states in each model to cluster the states
into similar states across models. We then compare the maximum abundance
of each state. 

This helps us visualize how stable the states are across models, and which
states are fragmented into other states.

```{r set_param}
presence.threshold = 0.3 #the minimum abundance of a histone mark to be considered present in the state
```

```{r source_code}
library("here")
all.code.dir <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.code.dir)){
	all.fun <- list.files(all.code.dir[i], full.names = TRUE, pattern = ".R")
	for(j in 1:length(all.fun)){source(all.fun[j])}
}
library(pheatmap)
```

```{r binary_states}
binary_states <- t(sapply(0:15, function(x) number2binary(x, 4)))
```

```{r find_state_fun}
find_state <- function(all.states, bin.state, thresh = 0.2){
    has.mark <- which(bin.state == 1)
    no.mark <- which(bin.state == 0)
    has.mark.idx <- Reduce("intersect", lapply(has.mark, function(x) which(all.states[,x] >= thresh)))
    no.mark.idx <- Reduce("intersect", lapply(no.mark, function(x) which(all.states[,x] < thresh)))
    if(length(has.mark) > 0 && length(no.mark) > 0){
        state.idx <- intersect(has.mark.idx, no.mark.idx)
    }
    if(length(has.mark) == 0){
        state.idx <- no.mark.idx
    }
    if(length(no.mark) == 0){
        state.idx <- has.mark.idx
    }
    #pheatmap(all.states[state.idx,])
    return(all.states[state.idx,,drop=FALSE])
}

```

```{r emissions}
histone.order <- c("H3K27me3", "H3K4me3", "H3K4me1", "H3K27ac")

all.models <- list.files(here("Data", "ChromHMM"))
state.num <- as.numeric(sapply(strsplit(all.models, "_"), function(x) x[1]))
model.order <- order(state.num, decreasing = FALSE)
ordered.models <- all.models[model.order]
emissions <- vector(mode = "list", length = length(all.models))
names(emissions) <- ordered.models
for(i in 1:length(ordered.models)){
    emissions[[i]] <- as.matrix(read.table(here("Data", "ChromHMM", ordered.models[i], 
    paste0("emissions_", state.num[model.order[i]], ".txt")), sep = "\t", header = TRUE,
    row.names = 1))
    #sort by histone modification
    mark.order <- match(histone.order, colnames(emissions[[i]]))
    #colnames(emissions[[i]])[mark.order]
    emissions[[i]] <- emissions[[i]][,mark.order]
}

#t(sapply(emissions, colnames))

all.states <- Reduce("rbind", emissions)
model.names <- unlist(lapply(1:length(emissions), 
    function(x) paste0("model_", rep(state.num[model.order[x]], 
    nrow(emissions[[x]])), "_state_", 1:nrow(emissions[[x]]))))
rownames(all.states) <- model.names
colnames(binary_states) <- colnames(all.states)

grouped.states <- lapply(1:nrow(binary_states), 
    function(x) find_state(all.states, binary_states[x,], presence.threshold))
```

## Grouped States {.tabset .tabset-fade .tabset-pills}

The following images show the states that were grouped together from different
models. These plots confirm the consistency of states from model to model.

```{r plot_state_groups, results = "asis", fig.height = 7, fig.width = 7}
for(i in 1:length(grouped.states)){
    cat("###", binary_states[i,], "\n")
    par(mar = c(4, 8, 2, 2))
    imageWithText(round(grouped.states[[i]], 2), global.color.scale = TRUE, 
        use.pheatmap.colors = TRUE, global.min = 0, global.max = 1, col.text.rotation = 0,
        col.text.adj = 0.5, row.text.shift = 0.2, cex = 1)
    cat("\n\n")
}
```

## State Abundance

```{r read_state_abundance}
state.abundance  <- lapply(ordered.models, 
    function(x) readRDS(here("Results", "ChromHMM", x, "State_Abundance.RDS")))
abundance.vector <- unlist(state.abundance)
grouped.abundance <- lapply(1:length(grouped.states), 
    function(x) abundance.vector[match(rownames(grouped.states[[x]]), model.names)])
for(i in 1:length(grouped.abundance)){
    names(grouped.abundance[[i]]) <- rownames(grouped.states[[i]])
}

#pheatmap(binary_states, cluster_rows = FALSE, cluster_cols = FALSE)
```

Build a grid that has the abundance for each state in each model.
Rows represent a single binary state. Columns represent a single model,
for example, the 4-state or 9-state model.

```{r state_grid, fig.width = 9, fig.height = 5}
#pheatmap(grouped.states[[15]], cluster_rows = FALSE, cluster_cols = FALSE)
#grouped.abundance[[15]]
abundance.grid <- matrix(NA, nrow = nrow(binary_states), ncol = length(ordered.models))
colnames(abundance.grid) <- sapply(strsplit(ordered.models, "_"), function(x) as.numeric(x[1]))

for(i in 1:nrow(binary_states)){
    if(length(grouped.abundance[[i]]) > 0){
        all.abundance <- grouped.abundance[[i]]
        model.size <- sapply(strsplit(names(all.abundance), "_"), function(x) as.numeric(x[2]))
        ordered.size <- as.character(sort(model.size, decreasing = FALSE))
        abundance.grid[i, ordered.size] <- all.abundance
    }
}

layout.mat <- matrix(1:3, nrow = 1)
layout(layout.mat, widths = c(0.2, 1, 0.1))
par(mar = c(4,2,2,0))
imageWithText(binary_states, global.color.scale = TRUE, global.max = 1, global.min = 0, 
use.pheatmap.colors = TRUE, cex = 1, col.text.shift = -1)
par(mar = c(4,0,2,2))
imageWithText(round(abundance.grid, 2), global.color.scale = TRUE, global.max = 1, global.min = 0, 
use.pheatmap.colors = TRUE, col.text.rotation = 0, cex = 1, col.text.shift = -0.5, col.text.adj = 0.5,
col.text.cex = 1.5)
par(mar = c(4,2,2,2))
imageWithTextColorbar(abundance.grid, global.color.scale = TRUE, global.max = 1, global.min = 0, 
use.pheatmap.colors = TRUE, cex = 1.5)

```

## State Effects

Also look at the state effects based on these groupings.
The following effects results were generated in 1.3_Compare_ChromHMM_States.Rmd

```{r state_effects, fig.width = 9, fig.height = 5}
state.effects <- readRDS(here("Results", "ChromHMM", "across.strain.models.RDS"))
effect.vector <- unlist(lapply(state.effects, function(x) x[,3]))
names(effect.vector) <- rownames(all.states)

grouped.effect <- vector(mode = "list", length = length(grouped.states))
for(i in 1:length(grouped.states)){
    effect.locale <- match(rownames(grouped.states[[i]]), names(effect.vector))
    grouped.effect[[i]] <- effect.vector[effect.locale]
}

effect.grid <- matrix(NA, nrow = nrow(binary_states), ncol = length(ordered.models))
colnames(effect.grid) <- sapply(strsplit(ordered.models, "_"), function(x) as.numeric(x[1]))

for(i in 1:nrow(binary_states)){
    if(length(grouped.effect[[i]]) > 0){
        all.effect <- grouped.effect[[i]]
        model.size <- sapply(strsplit(names(all.effect), "_"), function(x) as.numeric(x[2]))
        ordered.size <- as.character(sort(model.size, decreasing = FALSE))
        effect.grid[i, ordered.size] <- all.effect
    }
}

layout.mat <- matrix(1:3, nrow = 1)
layout(layout.mat, widths = c(0.2, 1, 0.1))
par(mar = c(4,2,2,0))
imageWithText(binary_states, global.color.scale = TRUE, global.max = 1, global.min = 0, 
use.pheatmap.colors = TRUE, cex = 1, col.text.shift = -1)
par(mar = c(4,0,2,2))
imageWithText(round(effect.grid, 2), use.pheatmap.colors = TRUE, 
col.text.rotation = 0, cex = 1, col.text.shift = -0.5, col.text.adj = 0.5,
col.text.cex = 1.5)
par(mar = c(4,2,2,2))
imageWithTextColorbar(effect.grid, use.pheatmap.colors = TRUE, cex = 1.5, axis.line = -2)

#plot(as.vector(abundance.grid), as.vector(effect.grid))
#abline(h = 0)
```

Write out a table of annotations to use in plotting.
```{r write_annotations}

annotations <- c("heterochromatin", "active enhancer", "intragenic enhancer", 
    "active enhancer", "unannotated", "active promoter", "unannotated", "active enhancer",
    "repressor", "unannotated", "unannotated", "unannotated", "bivalent promoter", "unannotated", 
    "poised enhancer", "unannotated")

descriptions <- c("strong downregulation", "weak upregulation esp. near TES", 
"weak upregulation", "upregulation", "weak downregulation near TSS", 
"upregulation near TSS", "weak downregulation near TSS", "strong upregulation", 
"strong downregulation", "no clear effect", "weak downregulation near TSS", 
"no clear effect", "weak downregulation near TSS", "no effect", "downregulation near TSS", 
"no clear effect")

annotation.table <- cbind(binary_states, annotations, descriptions)
write.table(annotation.table, here("Data", "support_files", "annotation.table.txt"),
    sep = "\t", quote = FALSE, row.names = FALSE)
```