---
title: "Imputation in the DO"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---


## Introduction

The purpose of this workflow is to compare the eQTL scans with imputed SNPs
to eQTL scans with chromatin states.

```{r set_param}
num.states = 9
is.interactive = FALSE
#is.interactive = TRUE
```

```{r load_libraries}
library(here)
```


```{r source_code}
all.code.dir <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.code.dir)){
	all.fun <- list.files(all.code.dir[i], full.names = TRUE)
	for(j in 1:length(all.fun)){source(all.fun[j])}
}
```

```{r load_other_lib}
needed.packages <- c("grid", "gprofiler2")
load_libraries(needed.packages)
```

```{r read_data}
gene.info <- readRDS(here("Data", "RNASeq", "RNASeq_gene_info.RData"))
strain.table <- as.matrix(read.delim(here("Data", "support_files", "strain.color.table.txt"),
comment.char = "!", header = FALSE))

expr.data <- readRDS(here("Data", "DOQTL", "DO_Expression.RDS")) 
expr <- expr.data$expr
covar <- expr.data$covar

chrom.mats <- readRDS(here("Results", "ChromHMM", paste0(num.states, "_states_C"), 
paste0("Chromatin_States_", num.states, "_full_gene_1000.RData")))

chrom.r2 <- readRDS(here("Results", "chQTL", "chQTL.R2.RDS"))
methyl.r2 <- readRDS(here("Results", "RRBS", "RRBS.R2.RDS"))
snp.r2 <- readRDS(here("Results", "SNPs", "SNP.R2.RDS"))
eqtl.r2 <- readRDS(here("Results", "eQTL", "eQTL.R2.RDS"))

inbred.expr <- readRDS(here("Data", "RNASeq", "Strain_Scaled_C_Expression.RData"))

snp.alleles <- readRDS(here("Results", "SNPs", "SNP.by.Position.RDS"))
methyl.mats <- readRDS(here("Results", "RRBS", "Aligned.Methyl.Mats.RDS"))
```

## r2 Score Distributions

The following plots show the distributions of r2 scores from different
modalities: Haplotypes, Imputed Chromatin State, Imputed SNPs, and 
imputed DNA methylation.

For the imputed r2 scores, we used the maximum r2 score calculated
for each transcript. 

```{r find_max_r2}
max.snp <- sapply(snp.r2, function(x) if(length(x) > 1){max(x, na.rm = TRUE)}else{NA})
max.methyl <- sapply(methyl.r2, function(x) if(!all(is.na(x))){max(x, na.rm = TRUE)}else{NA})
max.chrom <- sapply(chrom.r2, function(x) if(length(x) > 1){max(x[,1], na.rm = TRUE)}else{NA})
```


```{r r2_dist}
r2.list <- list("Haplotype" = eqtl.r2[,1], "Chromatin_State" = max.chrom,
    "SNP" = max.snp, "DNA_Methylation" = max.methyl)
if(is.interactive){quartz()}
boxplot(r2.list, ylab = "Variance Explained")
sapply(r2.list, summary)
```

The following plot shows the comparison between the maximum SNP r2
score, chromatin r2 score, haplotype r2 score and DNA methylation 
r2 score.

These are quite highly correlated. Across all genes, chromatin state does 
better than SNPs in explaining gene expression. There is more weight
above the line y=x in the middle plot. Although there are 
many cases for which an individual SNP has a higher r2 score than the 
chromatin state (points below the y=x line). 

There is also a set of genes for which the best SNP r2 score is 0
but the best chromatin r2 scores run the gamut up to about 100, with
haplotype r2 scores for these genes going up to 150.

```{r snp_v_chrom, fig.width = 9, fig.height = 9}
common.transcripts <- Reduce("intersect", list(names(chrom.r2), rownames(eqtl.r2), 
    names(methyl.r2), names(max.snp)))
common.chrom.idx <- match(common.transcripts, names(chrom.r2))
common.eqtl.idx <- match(common.transcripts, rownames(eqtl.r2))
common.methyl.idx <- match(common.transcripts, names(methyl.r2))
common.snp.idx <- match(common.transcripts, names(max.snp))

all.r2 <- cbind(eqtl.r2[common.eqtl.idx,1], max.chrom[common.chrom.idx],
    max.snp[common.snp.idx], max.methyl[common.methyl.idx])
colnames(all.r2) <- c("Haplotype_R2", "Chromatin_R2", "SNP_R2", "Methylation_R2")
all.r2[which(all.r2 == -Inf)] <- NA
if(is.interactive){quartz()}
pairs(all.r2)
```

```{r plot_comp, fig.width = 9, fig.height = 6}
r2.pairs <- pair.matrix(1:ncol(all.r2))
layout.mat <- matrix(c(1,2,4,3,5,6), nrow = 2, byrow = TRUE)
if(is.interactive){quartz(width = 9, height = 6)}
layout(layout.mat)

axis.min <- 0
axis.max <- 1

for(i in 1:nrow(r2.pairs)){
    idx1 <- r2.pairs[i,1]
    idx2 <- r2.pairs[i,2]
    plot.new()
    plot.window(xlim = c(axis.min, axis.max), ylim = c(axis.min, axis.max))
    abline(0,1, col = "lightblue", lwd = 3)
    points(all.r2[,idx1], all.r2[,idx2], pch = 16,col = "darkgray", cex = 0.7)
    axis(1);axis(2)    
    mtext(colnames(all.r2)[idx1], side = 1, line = 2.5)
    mtext(colnames(all.r2)[idx2], side = 2, line = 2.5)
    mtext(paste(colnames(all.r2)[idx1], "v.", colnames(all.r2)[idx2]), side = 3)
}
```

```{r retrieval_fun}
get_snp_scan <- function(gene.id){
    snp.idx <- which(names(snp.r2) == gene.id)
    if(length(snp.idx) == 0){
        return(NA)
    }else{
        return(snp.r2[[snp.idx]])
    }
}

get_chrom_mat <- function(gene.id){
    chrom.idx <- which(names(chrom.mats) == gene.id)
    return(chrom.mats[[chrom.idx]])
}

get_chrom_scan <- function(gene.id){
    chrom.idx <- which(names(chrom.r2) == gene.id) 
    if(length(chrom.idx) == 0){
    return(matrix(NA, ncol = 1))
    }
    return(chrom.r2[[chrom.idx]])
}

get_methyl_mat <- function(gene.id){
    gene.name <- gene.info[which(gene.info[,"ensembl_gene_id"] == gene.id)[1],"external_gene_name"]
    gene.locale <- which(names(methyl.mats) == gene.name)
    if(length(gene.locale) == 0){return(NULL)}
    gene.mat <- methyl.mats[[gene.locale]]
    return(gene.mat)
}

get_methyl_r2 <- function(gene.id){
    gene.locale <- which(names(methyl.r2) == gene.id)
    r2.scores <- methyl.r2[[gene.locale]]
}

get_expr_order <- function(gene.id){
    gene.idx <- which(names(inbred.expr) == gene.id)
    gene.expr <- inbred.expr[[gene.idx]]
    return(gene.expr[order(gene.expr),,drop=FALSE])
}

get_snp_alleles <- function(gene.id){
    gene.idx <- which(names(snp.alleles) == gene.id)
    if(length(gene.idx) == 0){
        return(NA)
    }
    snp.array <- snp.alleles[[gene.idx]]
    #flatten to letters by position for each strain
    allele.calls <- apply(snp.array, 1, function(x) apply(x, 1, function(y) which(y > 0)[1]))
    #reduce complexity to show only where variants differ from reference (B6)
    has.diff <- which(apply(allele.calls, 2, function(x) length(unique(x))) > 1)
    if(length(has.diff) == 0){return(NULL)}
    trimmed.alleles <- allele.calls[,has.diff,drop=FALSE]
    for(p in 1:ncol(trimmed.alleles)){
        trimmed.alleles[which(trimmed.alleles[,p] == unlist(trimmed.alleles[2,p])),p] <- NA
    }
    return(trimmed.alleles)
}

get_haplotype_r2 <- function(gene.id){
    gene.idx <- which(rownames(eqtl.r2) == gene.id)
    return(eqtl.r2[gene.idx,1])
}

plot_together <- function(gene.id){
        
    gene.info.idx <- which(gene.info[,1] == gene.id)[1]
    if(is.na(gene.info.idx)){
        return(NULL)
    }

    gene.name <- gene.info[gene.info.idx,"external_gene_name"]
    gene.strand <- gene.info[gene.info.idx,"strand"]
    if(gene.strand == 1){
        tss.pos <- gene.info[gene.info.idx,"start_position"]
        tes.pos <- gene.info[gene.info.idx,"end_position"]
    }else{
        tss.pos <- gene.info[gene.info.idx,"end_position"]
        tes.pos <- gene.info[gene.info.idx,"start_position"]
    }

    hap.r2 <- get_haplotype_r2(gene.id)

    if(length(hap.r2) == 0){
    return(NULL)
    }

    gene.snp.r2 <- get_snp_scan(gene.id)
    gene.snp.alleles <- get_snp_alleles(gene.id)
    if(length(gene.snp.alleles) > 1){
        rownames(gene.snp.alleles) <- strain.table[order.strains(rownames(gene.snp.alleles), 
        strain.table[,1], strain.table),6]
    }

    gene.chrom.r2 <- get_chrom_scan(gene.id)
    gene.chrom.mat <- get_chrom_mat(gene.id)
    
    gene.methyl.mat1 <- get_methyl_mat(gene.id)
    if(length(gene.methyl.mat1) > 0){
        #quartz();pheatmap(gene.methyl.mat1, cluster_rows = FALSE, cluster_cols = FALSE)
        gene.methyl.mat2 <- t(apply(gene.methyl.mat1, 1, function(x) bin.vector(x, c(0, 50, 100))))
        #quartz();plot(gene.methyl.mat1, gene.methyl.mat2)
        #quartz();pheatmap(gene.methyl.mat2, cluster_rows = FALSE, cluster_cols = FALSE)
        gene.methyl.mat3 <- gene.methyl.mat2[,apply(gene.methyl.mat2, 2, function(x) all(!is.na(x))),drop=FALSE]
        #quartz();pheatmap(gene.methyl.mat3, cluster_rows = FALSE, cluster_cols = FALSE)
        gene.methyl.r2 <- get_methyl_r2(gene.id)
    }else{
        gene.methyl.mat3 <- gene.methyl.mat1
        gene.methyl.r2 <- NULL
        }
    
    expr.order <- get_expr_order(gene.id)

    if(length(expr.order) > 1){
        strain.order <- rev(order.strains(rownames(expr.order), 
            colnames(gene.chrom.mat), strain.table))
        if(length(gene.snp.alleles) > 1){
            snp.strain.order <- rev(order.strains(rownames(expr.order), 
            rownames(gene.snp.alleles), strain.table))
            snp.strain.order <- snp.strain.order[which(!is.na(snp.strain.order))]
        }
        methyl.strain.order <- order.strains(rownames(expr.order), 
            rownames(gene.methyl.mat3), strain.table)
    }else{
        strain.order <- 1:ncol(gene.chrom.mat)
        snp.strain.order <- 1:nrow(gene.snp.alleles)
        methyl.strain.order <- 1:nrow(gene.methyl.mat3)
    }
    
    if(length(gene.snp.r2) > 0 && !is.na(gene.snp.r2)){
        snp.pos <- as.numeric(sapply(strsplit(names(gene.snp.r2), "_"), function(x) x[2]))
    }
    chrom.pos <- as.numeric(rownames(gene.chrom.r2))
    methyl.pos <- as.numeric(rownames(gene.methyl.r2))

    all.pos <- c(snp.pos, chrom.pos, methyl.pos)
    #boxplot(list(snp.pos, chrom.pos, methyl.pos, c(tss.pos, tes.pos)))
    global.max.pos <- max(all.pos)
    global.min.pos <- min(all.pos)
    
    global.max.r2 <- max(c(gene.chrom.r2[,1], gene.snp.r2, gene.methyl.r2, 
        hap.r2), na.rm = TRUE)

    if(!is.finite(global.max.r2)){
        return(NULL)
    }

    if(is.interactive){quartz(width = 8, height = 8)}
    layout.mat <- matrix(c(1:5, 0, 0, 6, 6, 0), ncol = 2, byrow = FALSE)
    layout(layout.mat, heights = c(1,0.5, 1, 1, 1), widths = c(1,0.4))
    
    par(mar = c(0,4,4,4))
    plot(x = snp.pos, gene.snp.r2, pch = "X", col = "#e41a1c", cex = 1.5,
        ylab = "r2 Score", axes = FALSE,
    xlim = c(global.min.pos, global.max.pos), ylim = c(0, global.max.r2))
    points(x = as.numeric(rownames(gene.chrom.r2)), gene.chrom.r2[,1], pch = "+", 
        col = "#377eb8", cex = 2)
    points(x = as.numeric(names(gene.methyl.r2)), gene.methyl.r2, pch = "O",
        col = "#4daf4a", cex = 1)
    abline(h = hap.r2, lty = 2)
    axis(2)

    par(xpd = NA)
    legend(x = global.max.pos, y = global.max.r2, pch = c("X", "+", "O"), 
        col = c("#e41a1c", "#377eb8", "#4daf4a"), legend = c("SNP", "Chrom", "Methyl"))
    par(xpd = TRUE)

    par(mar = c(0,4,0,4))
    plot.new()
    plot.window(xlim = c(global.min.pos, global.max.pos), ylim = c(0,1))
    arrows(x0 = tss.pos, y0 = 0.5, x1 = tes.pos)

    par(mar = c(1,4,0,4))
  
    plot.chrom.mat(t(gene.chrom.mat[,strain.order]), num.states = num.states, 
        xlim = c(global.min.pos, global.max.pos))

    par(mar = c(2,4,1,4))

    if(length(gene.snp.alleles) > 0 && ncol(gene.snp.alleles) > 1){
        plot.chrom.mat(gene.snp.alleles[snp.strain.order,], num.states = 4, 
        xlim = c(global.min.pos, global.max.pos), ylab = "SNPs", 
        state.labels = c("A", "C", "G", "T"))
    }
    if(length(gene.snp.alleles) > 0 && ncol(gene.snp.alleles) == 1){
        barplot(gene.snp.alleles[snp.strain.order,], horiz = TRUE, 
        las = 2, main = paste(round(as.numeric(colnames(gene.snp.alleles))/1e6, 2), "Mb"))
        legend.labels = paste0(1:4, "=", c("A", "C", "G", "T"))
        legend("topleft", legend = legend.labels)
    }
    if(length(gene.snp.alleles) == 0){
        plot.text("No SNPs")
    }

    if(length(gene.methyl.mat3) > 0){
    plot.methyl.mat(gene.methyl.mat3[methyl.strain.order,,drop=FALSE], 
        xlim = c(global.min.pos, global.max.pos))
    }else{
    plot.text("No Methylation Information")
    }

    par(mar = c(4,4,4,4))
    if(length(expr.order) > 1){
        bar.cols <- strain.table[order.strains(rownames(expr.order), strain.table[,1], 
        strain.table),3]
        barplot(expr.order[,1], horiz = TRUE, col = bar.cols, las = 2, 
            main = "Inbred Hepatocyte Expression")
    }else{
        plot.text("No Inbred Expression Data")
    }

    mtext(gene.name, side = 3, outer = TRUE, line = -2.5)
}
#plot_together(gene.id)
```


## Maximum Effect {.tabset .tabset-fade .tabset-pills}

The following plots show examples of genes for which the maximum effect is 
from one of the three non-haplotype modalities.

```{r max_r2, results = "asis", fig.width = 10, fig.height = 8}
n.sample = 10
high.r2 <- which(all.r2[,1] > 0.5)
max.r2 <- apply(all.r2[high.r2,2:4], 1, function(x) which.max(x))
max.counts <- lapply(1:3, function(x) which(max.r2 == x))
#sapply(max.counts, length)

for(i in 1:3){
    cat("###", colnames(all.r2)[i+1], "{.tabset .tabset-fade .tabset-pills}\n")
    sample.genes <- sample(which(max.r2 == i), n.sample)
    for(b in 1:length(sample.genes)){
        gene.id <- rownames(all.r2)[high.r2[sample.genes[b]]]
        gene.name <- gene.info[which(gene.info[,1] == gene.id)[1], "external_gene_name"]
        cat("####", gene.name, "\n")
        plot_together(gene.id)
        cat("\n\n")
    }
    cat("\n\n")
}
```


## Irf5

The following plot shows data for Irf5, which is has a very clean haplotype 
and chromatin story. Interestingly, there are no SNPs that we can attribute
to the haplotype or chromatin effect.

```{r test_genes, fig.width = 10, fig.height = 8}
gene.name <- "Irf5"
gene.id <- gene.info[which(gene.info[,"external_gene_name"] == gene.name)[1],1]
plot_together(gene.id)
```

There are many genes that have identical r2 scores for SNPs
and methylation. I checked a few below, and they are not in the
same positions. I gather that this is simply a case of the SNPs
and the DNA mathylation coincidentally having the same haplotype
pattern. It is not causal.

```{r rnd_ex, eval = FALSE}
#head(all.r2)
gene.name = "Pkd2"
gene.id <- gene.info[match(gene.name, gene.info[,"external_gene_name"]),"ensembl_gene_id"]
plot_together(gene.id)

cent_chrom <- readRDS(here("Results", "ChromHMM", "9_states_C", "Chromatin.States.Gene.Coords.RDS"))

get_state_switch <- function(chrom.mat, state1 = 7, state2 = 3){
    tss.locale <- get.nearest.pt(as.numeric(rownames(chrom.mat)), 0)
    has.state1 <- length(which(chrom.mat[tss.locale,] == state1))
    has.state2 <- length(which(chrom.mat[tss.locale,] == state2))
    if(has.state1 > 0 && has.state2 > 0){
        return(1)
    }else{
        return(0)
    }
}

example_73 <- sapply(cent_chrom, function(x) if(length(x) > 1){get_state_switch(x, 7, 3)}else{NA})

has.switch <- which(example_73 > 0)

pdf("~/Desktop/examples.pdf")
for(i in 1:length(has.switch)){
    print(i)
    plot_together(gene.id = names(has.switch)[i])
}
dev.off()


#potential examples: Irf5, Pkd2, Psma8, Gbp7, Lad1, Snhg11, Fam171b, Bmyc, Sorl1, 
# Rarres1, Mtnr1a, Capg, Adra2b, Kcp, CCnd1
```

## Conclusions

There are high correlations between haplotype r2 score, SNP r2 score,
and chromatin r2 score, but they are not 1:1.

There are positions without SNPs that have variation in chromatin.
There are positions with variation in SNPs that do not have variation in 
chromatin.

The haplotype always does better than either SNPs or chromatin state.
Chromatin state usually does better than SNPs, but not always.
