---
title: "DO data matching"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

I'm having trouble matching up data from different sources. DO eqtl LOD scores
from DO478_tpm_20360_cis_allele_coef.RData don't match the LOD scores I get when
I rescan data from Svenson_DO850_for_eQTL_veiwer_v0.Rdata.

Here I am poking around to see if I can figure out what is going on.

```{r}
library(here)
```

The files I am examining are the following:
    * DO478_tpm_20360_cis_allele_coef: contains a list of eQTL results, but 
    I'm not sure who generated them. 
    
    * Svenson_DO850_for_eQTL_veiwer_v0: I got these data from Dan. This is 
    a typical eQTL viewer data set with genoprobs, expression, covariates,
    etc.
    The dataset.expr data has an expression data set called expr and one
    called norm. The one called expr has rankz normalized values.
    The values in norm look funny. I'm not sure how they were normalize.
    Both expression data sets have 478 individuals and 11770 transcripts.
    There is another data set called dataset.expr.petr. The expr data set
    in this table has 478 individuals and 12067 transcripts. These have 
    also been rankZ normalized. The raw dataset in Petr's set has 46517 
    transcripts in TPM. The thresholds he used removed a lot of transcripts!
    There is yet another data set in here with only 384 individuals. I'm not
    sure what that's all about. 
    * _Svenson_DO478_normalized_genes_v1 is just expression data. It has
    478 individuals and 32879 transcripts. These transcripts are mean centered
    and normalized, but I'm not sure how they were normalized.

All of these data sets are a little bit different. Given that Petr's is
rankZ normalized and has more transcripts than the other rankZ normalize
set, I think it makes sense to use that. I will not use the previous eQTL
results (the cis_allele_coef) file, since I'm not sure where they came
from, and I want to make sure I'm scanning chromatin and genetics with the
same expression data.

```{r load_data}
do.results <- readRDS(here("Data", "DOQTL", "DO478_tpm_20360_cis_allele_coef.RData"))
dan.data <- load(here("Data", "DOQTL", "Svenson_DO850_for_eQTL_veiwer_v0.Rdata"))
svenson.do <- readRDS("/Users/atyler/Documents/Data/DO/RNASeq/_Svenson_DO478_normalized_genes_v1.rds")
```


```{r exploring}
boxplot(svenson.do[,1:10])

boxplot(dataset.expr.petr$expr[,1:10]) #rankZ normalized
boxplot(dataset.expr.petr$raw[,1:10]) #TPM
boxplot(dataset.expr$expr[,1:10]) #rankZ normalized
boxplot(dataset.expr$raw[,1:10]) #TPM
boxplot(dataset.expr$norm[,1:10]) #not rankZ normalized
```

```{r direct_compare}
compare.datasets <- function(expr1, expr2, n.to.test = 10){
    common.trans <- intersect(colnames(expr1), colnames(expr2))
    expr1.locale <- match(common.trans, colnames(expr1))
    expr2.locale <- match(common.trans, colnames(expr2))
    sub.expr1 <- expr1[,expr1.locale]
    sub.expr2 <- expr2[,expr2.locale]
    for(i in 1:n.to.test){
        sample.idx <- sample(length(common.trans), 1)
        plot(sub.expr1[,sample.idx], sub.expr2[,sample.idx])
    }
}


par(mfrow = c(3,3))
compare.datasets(dataset.expr$expr, dataset.expr.petr$expr, 9)

par(mfrow = c(3,3))
compare.datasets(dataset.expr.petr$raw, dataset.expr$raw,9)
```