---
title: "Running ChromHMM Starting With BAM files"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r load_libraries, echo = FALSE}
library(here)
```

```{r source_code}
all.code.dir <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.code.dir)){
	all.fun <- list.files(all.code.dir[i], full.names = TRUE, pattern = ".R")
	for(j in 1:length(all.fun)){source(all.fun[j])}
}
```

This is the first script in the series that analyzes epigenetic data from the Genetics of 
Epigenetics Project at JAX. This study is a survey of strain difference in histone 
modifications across the genome. The nine strains included in the study are the following:

+ DBA/2J and eight CC founders: 
+ 129S1/SvImJ
+ A/J
+ C57BL/6J
+ CAST/EiJ
+ NOD/ShiLtJ
+ NZO/HlLtJ
+ PWK/PhJ
+ WSB/EiJ

Hepatocytes were collected from female mice from each of the strains. Some of the mice 
were treated with dexamethasone to assay the effect of this treatment on gene expression 
and histone modifications. The following four histone marks were assayed in the hepatocytes:

+ H3K4me1
+ H3K4me3
+ H3K27me3
+ H3K27ac

This workflow allows selection of individual or combinations of marks for differential
analysis. 

It also runs either the control animals or treatment animals as specified below.

```{r select_mark}
c.or.t <- "T" #C for control, T for treatment.
all.marks <- c("H3K4me1", "H3K4me3", "H3K27me3", "H3K27ac")
selected.mark <- c("H3K4me1", "H3K4me3", "H3K27me3", "H3K27ac")
#selected.mark <- c("H3K4me1")
#selected.mark <- c("H3K4me3")
#selected.mark <- c("H3K27me3")
#selected.mark <- c("H3K27ac")
excluded <- setdiff(all.marks, selected.mark)
```

Reads were aligned to individualized genomes and coordinates were mapped back to C57BL/6J 
mm10.

File name structure for bed files is in the following structure:
   mark_strain_individual.bed

The individual component indicates whether the animal is a control (C) or was treated
   with dexamethasone (T)

For example here are bed files for H3K27me3 measured in B6 mice, Control individual 1 
   and Treatment individual 2:
        h3k27me3_B6_B6C1.bed
        h3k27me3_B6_B6T2.bed

This workflow uses ChromHMM [@ChromHMM] to identify patterns in histone marks in these 
strains. ChromHMM uses a hidden markov model to learn patterns of binary combinations 
in genomic data. For example, each of the histone marks in this data set can be either 
present or absent. They can also be present or absent in combinations. H3Kme1 and H3K27ac 
can be present together and may have a particular biochemical meaning. ChromHMM finds 
all the combinations that are present in the data set. These are different states that 
can be present in different positions along the genome, and they can be assigned functional 
meaning. H3K4me1 and H3K4me3 together, for example, may indicate a strong enhancer at that 
location. We can use these states to associate histone states with expression changes in 
the inbred lines. For example, PWK and CAST may have a strong enhancer state at a 
particular gene and also highly express that gene, while the rest of the strains may 
have no enhancer, or a repressor state and express the gene at a low level.

Determining the ChromHMM states is the first step in this analysis. This workflow performs
that process in the following steps:

1. Convert bam files to bed files
2. Binarize the bed files
3. Run ChromHMM

Any code chunks that take longer than a few seconds to run are marked with the time they took
on the Helix cluster at JAX.

## Convert bam files to bed files
The first step in this process is to convert the binary bam files to bed files. Each line 
of a bed file indicates the genomic position of a sequenced read for the histone mark 
listed in the file name. 

For example, if we are looking at H3K27ac_129C1.bed, the bed file contains coordinates 
for reads associated with the H3K27ac mark in control animal 1 from the 129 strain. The 
row contains the chromosome, start position, and stop position, followed by information 
about the assay and strand information. Only the chromosome, start, and stop are used by 
ChromHMM.

We need to convert the bam files for all individual samples to bed files.

First, we create symbolic links from all bam files to our own data directory.
All bam files are in the following directory:

```{r file_dir}
strain.bam.dir <- "/projects/carter-lab/hepatocytes/chipseq-alignments/output/bamfiles_B6Co"
chromhmm.dir <- "/projects/carter-lab/hepatocytes/ChromHMM"
```

This directory has the files for the B6 mice as well, even though there is a separate 
directory labeled as B6.

We create symbolic links for all bam files and their indices to the data directory for 
this workflow

```{r new_bam_dir}
new.bam.dir <- file.path(chromhmm.dir, "data", "bamfiles")
if(!file.exists(new.bam.dir)){system(paste("mkdir", new.bam.dir))}
```

Next we create the symbolic links.

```{r symlink_bam, eval = FALSE}
strain.bam <- get.files(path = strain.bam.dir, want = ".bam", 
dont.want = c("unmapped"), full.names = TRUE)

for(i in 1:length(strain.bam)){
   new.bam.file <- file.path(new.bam.dir, basename(strain.bam[i]))
   link.line <- paste0("ln -s ", strain.bam[i], " ", new.bam.file)
   system(link.line)
}
```

Next, we use bamtobed to generate bed files from the bam files. The output for this step 
will be stored in the following directory:

```{r bed_dir}
bed.dir <- file.path(chromhmm.dir, "data", "bedfiles")
if(!file.exists(bed.dir)){system(paste("mkdir", bed.dir))}
```

After running this script, you can run 1.1_run_bedtools to convert bam files to bed files.