---
title: Variation in histone configurations correlates with gene expression across nine inbred strains of mice
output: 
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
fontsize: 11pt
geometry: margin=1in
bst: genome_research
---

Anna L. Tyler\textsuperscript{1,$*$}, Catrina Spruce\textsuperscript{1,$*$}, 
Romy Kursawe\textsuperscript{2}, Annat Haber\textsuperscript{2}, 
Robyn L. Ball\textsuperscript{1}, Wendy A. Pitman\textsuperscript{1},
Alexander D. Fine\textsuperscript{1}, Narayanan Raghupathy\textsuperscript{1}, 
Michael Walker\textsuperscript{1}, Vivek M. Philip\textsuperscript{1},
Christopher L. Baker\textsuperscript{1}, J. Matthew Mahoney\textsuperscript{1},
Gary A. Churchill\textsuperscript{1}, Jennifer J. Trowbridge\textsuperscript{1},
Michael L. Stitzel\textsuperscript{2}, Kenneth Paigen\textsuperscript{1},
Petko M. Petkov\textsuperscript{1,$\dagger$}, Gregory W. Carter\textsuperscript{1,$\dagger$}

\textsuperscript{1} The Jackson Laboratory for Mammalian Genetics, 
  600 Main St. Bar Harbor, ME, 04609

\textsuperscript{2}The Jackson Laboratory for Genomic Medicine
    299 Farmington Ave, Farmington, CT 06032

$^*$ equal contribution
$^\dagger$ corresponding authors

Running title: Epigenetic variation in nine inbred mouse strains

\pagebreak

```{r num_states, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.pos = "!H", out.extra = "")

library(here)
library(psychometric)
param <- read.delim(here("Data", "support_files", "param.txt"), header = FALSE, 
    stringsAsFactors = FALSE, row.names = 1)
num.states = as.numeric(param["num_states",])
ordered.marks <- strsplit(param["ordered_marks",], ",")[[1]]
results.dir <- here("Results", "ChromHMM", paste0(num.states, "_states_C"))
data.dir <- here("Data", "ChromHMM", paste0(num.states, "_states_C"))

all.code.dir <- list.files(here("Code"), full.names = TRUE)
for(i in 1:length(all.code.dir)){
	all.fun <- list.files(all.code.dir[i], full.names = TRUE, pattern = ".R")
	for(j in 1:length(all.fun)){source(all.fun[j])}
}

#read in state order, which is established in Figures.Rmd
state.idx <- readRDS(file.path(results.dir, "state.idx.RDS"))

#use state.idx to order the states as they are in the Figures
annotations <- as.matrix(read.delim(here("Data", "support_files", "annotation.table.txt")))[state.idx,]
emissions <- as.matrix(read.delim(file.path(data.dir, paste0("emissions_", num.states, ".txt")), row.names = 1))[state.idx,]

#make sure all marks are in the same order
annotations[,1:4] <- annotations[,match(ordered.marks, colnames(annotations)[1:4])]
emissions <- emissions[,match(ordered.marks, colnames(emissions))]
```

```{r id_state, echo = FALSE}
id_state <- function(tag = NULL, state.emissions = NULL, annotation.table, emissions){
  if(!is.null(tag)){
    state.emissions <- annotation.table[which(annotation.table[,"annotations"] == tag),1:4]
  }
  thresholded.emissions <- emissions
  thresholded.emissions[which(emissions < 0.3)] <- 0
  thresholded.emissions[which(emissions >= 0.3)] <- 1
  state.num <- Reduce("intersect", lapply(1:length(state.emissions), function(x) which(thresholded.emissions[,x] == state.emissions[x])))
  return(state.num)
}

repressor.state <- id_state("polycomb repress.", annotation.table = annotations, emissions = emissions)
active.promoter.state <- id_state("active promoter", annotation.table = annotations, emissions = emissions)
nomark.state <- id_state("no marks", annotation.table = annotations, emissions = emissions)
poised.enhancer.state <- id_state(state.emissions = c(1,1,1,0), annotation.table = annotations, emissions = emissions)
bivalent.promoter.state <- id_state(state.emissions = c(1,1,0,0), annotation.table = annotations, emissions = emissions)
active.enhancer.state <- id_state(state.emissions = c(0,1,1,1), annotation.table = annotations, emissions = emissions)
unknown.state <- id_state(state.emissions = c(0,1,0,0), annotation.table = annotations, emissions = emissions)
weak.intragenic.enhancer <- id_state(state.emissions = c(0,0,1,0), annotation.table = annotations, emissions = emissions)
strong.intragenic.enhancer <- id_state(state.emissions = c(0,0,1,1), annotation.table = annotations, emissions = emissions)
tes.intragenic.enhancer <- id_state(state.emissions = c(0,0,0,1), annotation.table = annotations, emissions = emissions)
imputation.results <- readRDS(file.path(results.dir, "All.Variance.Explained.RDS"))

```

```{r summary_stats, echo = FALSE}
mean.var.exp <- round(colMeans(imputation.results, na.rm = TRUE), 2)
var.exp.cor <- round(cor(imputation.results, use = "pairwise.complete.obs"), 2)
comp.pairs <- pair.matrix(colnames(imputation.results))
r.ci <- round(t(sapply(1:nrow(comp.pairs), function(x) CIr(var.exp.cor[comp.pairs[x,1], comp.pairs[x,2]], 
  n  = length(which(!is.na(rowMeans(imputation.results[,c(comp.pairs[x,1], comp.pairs[x,2])])))), 
  level = 0.99))), 3)
var.ci.test <- apply(imputation.results, 2, function(x) t.test(x, conf.level = 0.99))
var.ci <- t(sapply(var.ci.test, function(x) round(x[[4]], 3)))
```


# Abstract

The diversity outbred (DO) mice and their inbred 
founders are widely used models of human disease. 
However, although the genetic diversity of these 
mice has been well documented, their epigenetic 
diversity has not. Epigenetic modifications, such 
as histone modifications and DNA methylation, are 
important regulators of gene expression, and as such 
are a critical mechanistic link between genotype and 
phenotype. Therefore, creating a map of epigenetic 
modifications in the DO mice and their founders is 
an important step toward understanding mechanisms 
of gene regulation and the link to disease in this 
widely used resource. To this end, we performed a 
strain survey of epigenetic modifications in hepatocytes 
of the DO founders. We surveyed four histone modifications 
(H3K4me1, H3K4me3, H3K27me3, and H3K27ac), and DNA 
methylation. We used ChromHMM to identify 14 chromatin 
states, each of which represented a distinct combination 
of the four histone modifications. We found that the 
epigenetic landscape was highly variable across the DO 
founders and was associated with variation in gene 
expression across strains. We found that epigenetic 
state imputed into a population of DO mice recapitulated 
the association with gene expression seen in the founders 
suggesting that both histone modifications and DNA methylation 
are highly heritable mechanisms of gene expression regulation. 
We illustrate how DO gene expression can be aligned with inbred 
epigenetic states to identify putative cis-regulatory regions. 
Finally, we provide a data resource that documents strain-specific 
variation in chromatin state and DNA methylation in hepatocytes 
across nine widely used strains of laboratory mice.

# Introduction

The development of the diversity outbred (DO) mice 
\citep{Svenson:2012hq, churchill2012diversity, 
koyuncu2021cxcl1, kurtz2020diversity, bogue2015collaborative, 
kebede2014insights, keller2019gene} and their sister population, 
the collaborative cross (CC) \citep{threadgill2011collaborative, 
threadgill2012ten, durrant2011collaborative, mao2015identification, 
graham2021baseline}, has demonstrated the critical importance of 
genetic diversity in our understanding of disease biology.
These mice have been used to investigate the genetic architecture 
of complex disease \citep{pmid28592500}, to identify genetic 
modifiers of Mendelian disease \citep{takemon2021uncovering}, 
and to study the effects of genetic variation on susceptibility 
to infectious disease \citep{kurtz2020diversity}. These models
have huge potential to uncover mechanistic insights into 
multiple aspects of human health and disease. However, although
the genetic diversity of these mice is well documented, the
epigenetic diversity of these mice is relatively unknown. 
This gap in knowledge is potentially critical to understanding 
mechanisms of gene regulation and its impact on disease in 
these important models. 

Epigenetic modifications, such as histone modifications 
\citep{pmid32671792, pmid29625185} and DNA methylation 
\citep{pmid21701563, pmid20720541}, regulate gene 
expression by modifying the accessibility of DNA to
transcription machinery \citep{lawrence2016lateral, 
jones2012functions, moore2013dna}. These modifications 
allow an organism to develop an array of diverse cell 
types from a single genome. 

Epigenetic modifications also vary across individuals.
Studies in humans \citep{mcvicker2013identification, kang2021genetic}, 
rats \citep{rintisch2014natural}, and cattle 
\citep{prowse2022genetic} have identified genetic loci
associated with variation in histone modification.
Heritable variation in histone modification has also 
been linked to variation in gene expression 
\citep{kang2021genetic, rintisch2014natural, prowse2022genetic} 
and clinical outcomes \citep{kang2021genetic}. Similar studies
have associated heritable variation in DNA methylation
to clinical outcomes, such as immune dysregulation
\citep{hawe2022genetic}.

This link between genetic variation and epigenetic 
variation may be a critical piece of understanding 
genetic susceptibility to disease. Because the majority 
of disease-associated genetic variants discovered in 
humans are in gene regulatory regions, it has been 
suggested that it is the regulation of gene expression, 
rather than alteration of protein function, that is the 
primary mechanism through which genetic variants confer 
disease risk \citep{maurano2012systematic, farh2015genetic, pmid21617055, pmid19474294}.
Detailed epigenomic landscapes, therefore, may provide 
important mechanistic insight connecting genotype to 
disease risk.

To address the lack of documented epigenetic variation
in the DO/CC founders, we undertook a strain survey of 
epigenetic variation in hepatocytes across the eight 
founders of the DO/CC mice, as well as DBA/2J, which, 
along with C57BL/6J, is one of the founders of the widely 
used BxD recombinant inbred panel of mice \citep{Ashbrook:2019bd}.

We assayed four histone modifications: H3K4me3, which is 
associated with promoter regions 
\citep{heintzman2007distinct, pmid15680324}, H3K4me1, which 
is associated with enhancer regions \citep{heintzman2007distinct},
H3K27me3, which is associated with polycomb repression 
\citep{bonasio2010molecular}, and H3K27ac, which has been 
associated with active enhancers and promoters 
\citep{creyghton2010histone, heintzman2009histone, rada2011unique}.
We also assayed DNA methylation which is differentially 
associated with gene expression depending on its position
relative to the gene \citep{moore2013dna, jones2012functions}.
Methylation of DNA in promoters inactivates the promoters
thereby reducing gene expression, whereas methylation of DNA 
in insulators inactivates the insulators thereby increasing 
expression of the targeted gene \citep{jones2012functions}.

We used ChromHMM \citep{Ernst:2012ii} to identify `r num.states` 
chromatin states, each representing a unique combination of 
the four histone marks. We investigated the association between 
variation in these epigenetic markers and variation in gene 
expression across the nine inbred strains. 

We extended our analysis into a population of DO mice to 
investigate the heritability of histone modifications and
DNA methylation with respect to gene expression. To do this,
we imputed the `r num.states` chromatin states and DNA methylation 
into the DO mice. We then mapped gene expression to the imputed 
epigenetic states to assess the extent to which expression
quantitative trait loci (eQTLs) in the DO were related to 
variation in epigenetic modifications. 

# Results

Both gene expression and epigenetic state were consistent 
within each inbred mouse strain but varied across the 
strains suggesting strong genetic regulation of both 
modalities. This is seen as a clustering of individuals 
from the same strain in principal component plots of 
transcriptomic and epigenetic features (Fig. \ref{fig:pc_plots}). 
Patterns of gene expression (Fig. \ref{fig:pc_plots}A), 
DNA methylation (Fig. \ref{fig:pc_plots}B) and individual 
histone modifications (Fig. \ref{fig:pc_plots}C-F) clustered 
in similar patterns, although a relatively small percent of 
the variation in the methylome was related to strain. The 
three subspecies \textit{musculus} (in red), \textit{castaneous} 
(in green) and \textit{domesticus} (all others) were widely 
separated suggesting that subspecies structure made up the 
majority of the observed variance. The \text{domesticus} 
strains largely clustered together. These data provide evidence 
that epigenetic features relate to gene expression in a manner 
that is consistent with the subspecific origin of the mouse 
strains \citep{yang2007subspecific}. For a more detailed 
visualization of the correlations between strains see Supplemental 
Figure \ref{supp_fig:feature_correlation}. Also, note that 
all genes used in this analysis were expressed at a minimal
level across the strains (overall mean of 5 TPM), so results
do not include data from non-expressed genes.

\begin{figure}[ht!]
\includegraphics[width=3in]{Figures/Fig1_Decomposition.png} 
\caption{The first two principal components of each genomic 
feature across nine inbred mouse strains. In all panels each 
point represents an individual mouse, and strain is indicated 
by color as shown in the legend at the bottom of the figure. 
Three individuals per strain are shown. Each panel is labeled 
with the data used to generate the PC plot. (A) Hepatocyte 
transcriptome - all transcripts expressed in isolated hepatocytes. 
(B) DNA methylation - the percent methylation at all CpG sites 
shared across all individuals. (C-F) Histone modifications - 
the peak heights of the indicated histone modification for
positions aligned across strains.}
\label{fig:pc_plots}
\end{figure}

## Chromatin state overview

We used ChromHMM to identify `r num.states` chromatin states 
composed of unique combinations of the four histone modifications
(Fig. \ref{fig:state_overview}A). We calculated the enrichment 
of each state near predicted functional elements in the mouse
liver (Fig. \ref{fig:state_overview}B), and correlated the
presence of each state with gene expression both across genes
and across the inbred strains (Fig. \ref{fig:state_overview}C).

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Fig2_Overview.png} 
\caption{Overview of chromatin state composition, genomic 
distribution, and association with expression. (A) Emission 
probabilities for each histone modification in each 
chromatin state. Blue indicates the absence of the histone
modification, and red indicates the presence of the 
modification. (B) The distribution of each state around 
functional elements in the genome. Red indicates that the 
state is enriched near the annotated functional element. 
Blue indicates that the state is depleted near the annotated 
functional element. Abbreviations are as follows: 
Enh. = enhancer, Tsd = distal to the transcription start site, 
Tsp = proximal to the transcription start site; Hetero. = 
heterochromatin; FR = flanking region. (C) The association between 
chromatin state variation and gene expression. Bars are colored 
based on the size and direction the state's association with expression. 
Red/blue bars show the associations of chromatin state with gene
expression across strains. Blue-gray bars show the associations of
chromatin state with gene expression across genes. (D) Plausible 
annotations for each state based on genomic enrichments and association
with gene expression. The numbers in parentheses indicate the percent 
of the genome that was assigned to each state. Repress. = repressor.}
\label{fig:state_overview}
\end{figure}

To associate chromatin state with expression across transcripts
(Fig. \ref{fig:state_overview}C blue-gray bars), we calculated the 
proportion of each gene body that was occupied by each state in each 
inbred strain. We then fit a linear model to associate the proportion 
of each chromatin state with the amount of transcription (Methods). 
We did this separately in each strain. Some chromatin states, such 
as State `r active.enhancer.state` were more abundant in highly 
expressed genes, whereas other states, such as State `r repressor.state`, 
were more abundant in lowly expressed genes.

We compared this correlation to the correlation between chromatin 
state and gene expression across strains (Fig. \ref{fig:state_overview}C
red/blue bars) (Methods). To do this, we normalized the expression of 
each transcript and the proportion of each chromatin state across 
strains (Methods). We then fit a linear model to estimate whether 
the proportion of each state varying across strains was associated 
with gene expression. For any given transcript, strains with greater 
proportions of State `r active.enhancer.state` had higher expression 
than strains with lower proportions of State `r active.enhancer.state`. 
Through this calculation, we can associate strain variation in chromatin 
state with strain variation in gene expression. 

In Figure \ref{fig:state_overview}, the states are ordered by their
association with gene expression across strains, which helps illustrate 
several patterns. Overall, states that were associated with increased 
expression across transcripts were also associated with increased 
expression when varying across strains. The state with the largest 
negative association with gene expression across strains, 
State `r nomark.state`, was the absence of all measured modifications. 
Other states associated with reduced gene expression contained the 
repressive mark H3K27me3. The states with the largest positive correlations
with gene expression all had some combination of the activating marks H3K4me3, 
H3K4me1, and H3K27ac. The repressive mark was less commonly seen in 
these activating states. 

We used the functional element enrichments to assign putative 
annotations to each of the `r num.states` chromatin states 
(Fig. \ref{fig:state_overview}D). Except for State `r nomark.state`, 
all states were enriched around at least one of the predicted 
functional elements in mouse liver (Fig. \ref{fig:state_overview}B). 
Where there was more than one obvious enrichment for the state, 
we used our own associations 
with gene expression to narrow down which regulatory label we 
assigned each state. Interestingly, the enrichments of these states 
largely matched the associations we saw between each state and 
gene expression (Fig. \ref{fig:state_overview}C). For example, 
State `r active.enhancer.state`, which was enriched around strong 
enhancers, was the state that was most strongly correlated with 
increased expression both across genes, and across strains. Likewise, 
States 2-4 were all enriched around active enhancers or promoters, 
and were all correlated with increased expression overall. 

At the other end of the spectrum, state `r repressor.state` was 
enriched around polycomb repressor marks, as we would expect 
because it was defined by presence of H3K27me3, which is associated 
with polycomb repression. This state was also correlated with reduced
expression both across genes and across strains. 

Many of the states with weaker associations with gene expression, 
both positive and negative were most enriched around bivalent promoters. 
This suggests that the bivalent promoter class may represent a diverse 
array of functional elements with varied effects on gene expression, and 
that more detailed experiments investigating the relationship between 
these states and gene expression could potentially identify novel 
chromatin states influencing expression in these cells. 

## DNA methylation overview

To investigate the variation in DNA methylation across the
inbred strains, we examined both strain-specific CpG 
sites and strain-specific methylation values. We defined a 
strain-specific CpG site as one that was present in all 
individuals in at least one strain and absent in all 
individuals in at least one other strain.

Roughly 17.8\% of all CpG sites were strain-specific 
ranging from 16\% to 19\% across the chromosomes. 
Strain-specific CpG sites were more commonly present in 
CAST, PWK, and B6 compared to the other strains (Fig. 
\ref{fig:cpg_overview}A), suggesting that these strains 
have more DNA methylation overall compared with the other 
strains.

\begin{figure}[ht!]
\begin{minipage}[c]{0.4\textwidth}
  \includegraphics[width=\textwidth]{Figures/Fig3_CpG_Overview.png} 
\end{minipage}\hfill
\begin{minipage}[c]{0.5\textwidth}
\caption{Overview of strain-specific CpG sites. (A) Boxes 
show proportion of strain-specific CpG sites that are present 
in each strain. Boxes are colored by official strain colors
for ease of visualization. Short names for strains are
indicated below each box. (B) The $log_{10}$(Fold Enrichment)
of CpG sites shared across all strains (green) and those
that are strain-specific (purple). (C) A comparison
of enrichments between CpG sites that are shared across all 
strains and those that are strain-specific. Bars above 1
show where strain-specific CpGs were more enriched than 
shared CpGs. Bars below 1 indicate where strain-specific 
CpGs were less enriched than shared CpGs. The vertical 
line marks where shared and strain-specific CpGs were equally 
enriched. Abbreviations are as follows: FR - flanking region; 
Tsp - transcription start site proximal; Tsd - transcription
start site distal, Hetero. - heterochromatin; Enh. - enhancer.
}
\label{fig:cpg_overview}
\end{minipage}
\end{figure}


CpG sites that were shared across all strains were enriched
around genomic features such as CpG islands and promoters 
(Methods) (Fig. \ref{fig:cpg_overview}B green). Strain-specific 
CpG sites were also enriched around CpG islands 
and promoters (Fig. \ref{fig:cpg_overview}B purple). 
However, relative to the CpG sites found in all strains, 
the strain-specific CpG sites were more strongly enriched 
specifically in enhancers, especially TSS-distal poised enhancers 
and weak enhancers (Fig. \ref{fig:cpg_overview}C). Relative to 
the CpG sites common across all strains, strain-specific sites 
were depleted in promoter regions and CpG islands (Fig. 
\ref{fig:cpg_overview}C) suggesting that variation in DNA 
methylation across strains primarily occurs in enhancers that 
fine-tune gene expression levels rather than in promoters which 
might result in genes being turned on or off.

## Spatial distribution of epigenetic modifications around gene bodies

In addition to looking for enrichment of chromatin states and
CpG sites near annotated functional elements, we characterized 
the fine-grained spatial distribution of these features around 
gene bodies by normalizing genomic positions to run from 0 at 
the TSS to 1 at the TES (See Methods) (Fig. \ref{fig:state_abundance})

\begin{figure}[ht!]
\includegraphics[width=0.67\textwidth]{Figures/Fig4_Abundance.png} 
\caption{Relative abundance of chromatin states and methylated DNA. (A) Each panel 
shows the abundance of a single chromatin state relative to gene TSS and TES. The 
$y$-axis in each panel is the proportion of genes containing the state. Each
panel has an independent $y$-axis to better show the shape of each curve.
The $x$-axis is the relative gene position. The TSS and TES are marked as vertical
gray dashed lines. (B) The same data shown in panel A, but with all states overlayed
onto a single set of axes to show the relative abundance of the states. 
(C) The density of CpG sites relative to the gene body. The $y$-axis shows 
the inverse inter-CpG distance in base pairs. The density is highest near the 
TSS. CpG sites are less dense within the gene body and in the intergenic space. 
(D) Percent methylation relative to the gene body. The $y$-axis shows the median 
percent methylation at CpG sites, and the $x$-axis shows relative gene position. 
CpG sites near the TSS are unmethylated relative to intragenic sites and to sites
just upstream and downstream of the gene bodies. In both C and D standard error 
is shown as a blue envelope around the mean; however, the standard error is so 
small that it is not visible in the figure.}
\label{fig:state_abundance}
\end{figure}

The spatial patterns of the individual chromatin states are shown 
in (Fig. \ref{fig:state_abundance}A), and an overlay of all 
states together (Fig. \ref{fig:state_abundance}B) emphasizes the
difference in abundance between the most abundant states (States 
`r active.enhancer.state`, `r active.promoter.state`, and 
`r nomark.state`), and the remaining states, which were 
relatively rare.

Each chromatin state had a characteristic distribution pattern
across the gene body. For example, State `r nomark.state`, 
which was characterized by the absence of all measured histone 
modifications, was strongly depleted near the TSS, indicating 
that this region is commonly subject to the histone modifications 
we measured here. It should be noted that this pattern is independent
of the global enrichment patterns shown in Figure \ref{fig:state_overview}. 
Although state `r nomark.state` is generally depleted in gene bodies relative
to intergenic regions, it is especially depleted at the TSS.
In contrast, States `r active.enhancer.state` and `r active.promoter.state` 
were both relatively abundant at the TSS. State `r active.promoter.state` 
was very narrowly concentrated right at the TSS, consistent with
its annotation as an active promoter (Fig. \ref{fig:state_overview}). 
State `r active.enhancer.state` on the other hand, was especially 
enriched just upstream of the TSS, consistent with its annotation 
of a TSS-proximal strong enhancer. State `r strong.intragenic.enhancer`, 
was depleted near the TSS, but enriched within the gene body, consistent 
with its annotation of a TSS-distal enhancer.

States with weaker associations to expression (indicated by grayer
shades in Fig. \ref{fig:state_abundance}) were of lower abundance, 
but had distinct distribution patterns around the gene body suggesting 
the possibility of distinct functional roles in the regulation of gene 
expression. The abundance patterns were identical across all strains 
(Supplemental Fig. \ref{supp_fig:state_abundance_by_strain})

DNA methylation showed similar characteristic variation in abundance
(Fig. \ref{fig:state_abundance}C-D). The TSS had densely packed 
CpG sites relative to the gene body (Fig. 
\ref{fig:state_abundance}C). As expected, the median CpG site near 
the TSS was consistently hypomethylated relative to the median CpG 
site (Fig. \ref{fig:state_abundance}D). All genes used in this analysis 
were expressed and thus had some degree of hypomethylation. There were 
also no large-scale differences in CpG distribution or percent methylation 
across strains (Supplemental Fig. \ref{supp_fig:percent_methyl_by_strain}).

## Spatially resolved associations with gene expression

The distinct spatial distributions of the chromatin states and 
methylated CpG sites around the gene body raised the question 
as to whether the associations of these states with gene expression 
could also be spatially resolved. To investigate this possibility 
we tested the association between gene expression and both 
chromatin state and DNA methylation using spatially resolved models
(Methods). We tested the association of each chromatin state with 
expression across genes within hepatocytes (Fig. 
\ref{fig:state_effects} left column) and the association of each 
chromatin state with the variation in gene expression across 
strains (Fig. \ref{fig:state_effects} middle column).

\begin{figure}[ht!]
\includegraphics[width=0.7\textwidth]{Figures/Fig5_Effects.png}
\caption{Associations of chromatin states with gene expression. Each 
column shows the association of each chromatin state with gene 
expression in a different experimental context as labeled. 
Effects shown are $\beta$ coefficients from equation 1. The 
$y$-axes vary across each row to emphasize the shape of each effect, 
so $y$-axis labels indicate only positive and negative effects. 
Colored areas show the 95\% confidence interval around each 
estimate. The final column shows the annotation of each state for 
comparison with its association with gene expression. All $x$-axes 
show the relative position along the gene body running from just 
upstream of the TSS to just downstream of the TES. Vertical gray 
dashed lines mark the TSS and TES in all panels.
}
\label{fig:state_effects}
\end{figure}

All chromatin states demonstrated spatially dependent associations 
with gene expression within hepatocytes. Figure \ref{fig:state_effects}
shows how these associations are distributed across the states 
and across the gene bodies. For many of the states, the associations 
with expression were concentrated at or near the TSS, while in the other 
states associations were seen across the whole gene. The direction of 
the coefficients matched the overall associations of each state seen 
previously (Fig. \ref{fig:state_overview}), but here we see the effects 
in finer resolution. For example, State `r active.promoter.state` was 
positively correlated overall with gene expression (Fig. 
\ref{fig:state_overview}C), but in Figure \ref{fig:state_effects} 
we see that this positive correlation is primarily limited to the 
region near the TSS, consistent with its annotation as a promoter state.

Further, the spatial associations observed across genes 
(Fig. \ref{fig:state_effects} left column) were largely recapitulated 
in the measurements across strains (Fig. \ref{fig:state_effects} middle 
column). That is, chromatin states that either enhanced or suppressed gene 
expression across hepatocyte genes were similarly related to variation in 
expression across strains. This suggests that the genetic differences 
between strains modify chromatin activity in a manner similar to that 
used across genes. One notable exception was State 
`r weak.intragenic.enhancer`, whose presence upregulated genes within 
hepatocytes, but was not associated with expression variation across strains.

We also examined the association of percent DNA methylation with 
gene expression across genes and across strains (Fig. 
\ref{fig:DNA_methylation_effect}). As expected, methylation at 
the TSS was associated with lower expressed genes in hepatocytes
(Fig. \ref{fig:DNA_methylation_effect}A). 
We did not detect an association between DNA methylation percent
and gene expression across inbred strains, perhaps because there
were too few strains to reliably estimate the coefficients 
(Fig. \ref{fig:DNA_methylation_effect}B).

\begin{figure}[ht!]
\includegraphics[width=3in]{Figures/Fig6_methylation_effect.png} 
\caption{Association of DNA methylation wkth gene expression (A) across 
gene expression in hepatocytes and (B) across inbred strains. The dark 
gray line shows the estimated effect of percent DNA methylation on gene 
expression. The $x$-axis is normalized position along the gene body 
running from the transcription start site (TSS) to the transcription 
end site (TES), marked with vertical gray dashed lines. The horizontal 
solid black line indicates an association of 0. The shaded gray area shows 
95\% confidence interval around the model fit.}
\label{fig:DNA_methylation_effect}
\end{figure}


## Interactions between chromatin state and DNA methylation

We investigated whether there was an interaction between DNA
methylation and chromatin state by asking two questions. First,
were CpG sites within different chromatin states methylated at
different levels? And second, was DNA methylation within specific
chromatin states differentially associated with gene 
expression across inbred mice? If DNA methylation essentially
inactivates a region of DNA, methylation in a region identified
as a repressor based on its chromatin state might be expected 
to increase gene expression, whereas methylation in an active 
enhancer might decrease gene expression. 

To investigate these questions, we identified CpG sites within 
each of the `r num.states` chromatin states. We calculated the 
average percent methylation of these sites, and the association of 
DNA methylation with gene expression for each set of sites (Methods). 
We treated missing CpG sites in individual strains as unmethylated.

Although methylation patterns in all states followed roughly the same
pattern of being unmethylated at the TSS and methylated within the gene 
body, values ranged widely across the states from State 
`r active.promoter.state` with a mean of 27\% methylated DNA intragenically,
to State `r nomark.state` with 83\% methylated DNA intragenically. Again,
these differential levels of methylation within these states are 
consistent with the state annotations. State `r active.promoter.state`
was annotated as an active promoter, and we would expect DNA methylation
in this state to be low. State `r nomark.state` has no histone modifications
and is not expected to be transcriptionally active, which is consistent
with high levels of DNA methylation.

DNA methylation within each chromatin state was differentially
correlated with gene expression (Supplemental Fig. 
\ref{supp_fig:methyl_effect_by_state}). DNA methylation in State 
`r active.promoter.state`, the active promoter state, was associated 
with decreased gene expression, suggesting that DNA methylation in 
this state deactivated the active promoter state. Overall, the
repressor state, State `r repressor.state`, was negatively associated 
with gene expression. However, DNA methylation in this state was 
positively associated with gene expression, suggesting that this 
repressive state can be inactivated by DNA methylation.

## Imputed chromatin state was associated with gene expression in DO mice

Thus far, we have shown correlations between gene expression and 
epigenetic features in inbred mice. We were also interested in 
whether chromatin state and DNA methylation were associated with 
gene expression in an outbred mouse population. Although we did not 
measure epigenetic modifications directly in an outbred population,
we had liver gene expression from a previously published population 
of diversity outbred mice \citep{pmid28592500}. Inheritance of chromatin
state and DNA methylation is complex \citep{rintisch2014natural}; 
however there is evidence that the heritability for both epigenetic 
features is high \citep{pmid16009939,pmid33931130} suggesting the 
possibility of imputing epigenetic features from local genotype into
the DO mice. Even with imperfect estimates of epigenetic features in 
the outbred mice, a common pattern of association between outbred and
inbred mice would support the idea that inherited variance in epigenetic 
features contributes to inherited variation in gene expression across 
genetically distinct individuals.

We imputed chromatin state, DNA methylation, and SNPs into the DO 
population (Methods). Because any feature imputed from haplotype 
will be correlated with anything that haplotype is correlated with, 
we performed permutations that shuffled the relationship between 
haplotype and chromatin state (Methods). The resulting $p$-value 
distributions of each genomic feature suggested that each imputed 
feature was significantly associated with gene expression in the DO 
beyond the effects of the imputation alone 
(Supp. Fig. \ref{supp_fig:empirical_p}).

We then tested the association between each imputed chromatin state, SNP,
or CpG site with gene expression in the DO. We tested each state
independently. This is not the standard method for testing associations 
between markers and gene expression in this type of population, but it 
constrains the degrees of freedom for all features to one so we can 
compare the associations across all features.

Haplotypes are a measure of ancestry, whereas chromatin state, DNA 
methylation, and SNPs are all potentially functional variation in 
the genome or epigenome. Thus any individual chromatin state, SNP, 
or CpG site can theoretically explain more variance in gene expression 
than an individual haplotype. 

Figure \ref{fig:effect_distrubutions} compares the variance 
explained by individual haplotypes with that explained by 
any individual chromatin state, CpG site, or SNP. All imputed
features--individual chromatin states (mean 14\%), DNA methylation 
(mean 14\%), and SNPs (mean 13\%)--explained more variance
in gene expression than individual haplotypes (mean 11\%) 
(Fig. \ref{fig:effect_distrubutions}A). This suggests that any 
given chromatin state, CpG site, or SNP carries more functional 
information than any individual haplotype, which is primarily a 
measurement of ancestry.

Figure \ref{fig:effect_distrubutions}B shows the maximum 
variance explained by each genomic feature for each transcript 
in the transcriptome. Dots above the line indicate transcripts 
for which the imputed genomic feature explained more variance 
than haplotype. Dots below the line indicate transcripts for 
which the imputed genomic feature explained less variance than 
haplotype. Individual haplotype explained less variance than 
any other genomic feature for the majority of transcripts. 
supporting the hypothesis that all these features carry 
heritable information that potentially regulates gene 
expression in this genetically diverse population. 

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Fig7_Imputation.png} 
\caption{Gene expression variance in a DO population explained 
by chromatin state compared to three other genomic features: 
local haplotype, local SNP genotype, and local imputed DNA 
methylation status. A. Distributions of gene expression variance 
explained by each feature. B. Direct comparisons of 
variance explained by local haplotype, and each of the other 
genomic features. Blue lines show $y = x$. Each point is a 
single transcript.}
\label{fig:effect_distrubutions}
\end{figure}

In addition to calculating overall associations, we calculated 
position-based associations between each epigenetic feature 
and gene expression (Fig. \ref{fig:state_effects} right column,
and Fig. \ref{fig:DNA_methylation_effect}C). The associations
in the DO mice largely matched those seen in the inbred mice
for both chromatin state and DNA methylation. Interestingly, 
even though DNA methylation showed no association with gene
expression across strains in the inbred mice, there was a
weak, but significant association with gene expression in the
DO mice. This may be due to the increased power to detect
effects in the 378 DO mice relative to the 9 inbred strains. 

## Hypothesis generation for cis-regulatory regions

By aligning associations with gene expression from the
DO mice with inbred epigenetic features, we can generate 
hypotheses about heritable cis-regulatory regions in 
these mice. In particular, for any gene whose variance 
was explained at least as well by an imputed feature as 
by haplotype, there is the possibility that the imputed 
feature marks a cis-regulatory element. This occurrence 
provides an opportunity to annotate novel functional 
elements in the mouse genome, or provide supportive
evidence of previously predicted functional elements. 
As an example, we investigated the gene \textit{Pkd2} 
(Fig. \ref{fig:example_gene}). 

WSB and PWK were low-expressing strains for \textit{Pkd2}, 
and the remaining strains had higher expression 
(Fig. \ref{fig:example_gene}E). The haplotype effects
in the DO mirror this pattern with the CAST allele showing an
especially high association with increased gene expression. 
Figure \ref{fig:example_gene}B, C, and D show chromatin state,
SNP genotype, and DNA methylation state along the body of \textit{Pkd2} 
respectively. Panel A shows the association of each of the
imputed features with gene expression in the DO. The detailed 
view of this gene identifies two particularly interesting regions 
marked by gray arrows in panel A. One is at the TSS and the 
immediately surrounding area, and the other is just downstream 
of the TSS. 

\begin{figure}[ht!]
\includegraphics[width=0.7\textwidth]{Figures/Fig8_Example.png} 
\caption{Example of epigenetic states and imputation results for a single 
gene, \textit{Pkd2}. The legend for each panel is displayed to its right.
(A) The variance in DO gene expression explained at 
each position along the gene body by each of the imputed genomic 
features: SNPs - red X's, Chromatin State - blue plus signs, and 
Percent Methylation - green circles. The horizontal 
dashed line shows the maximum variance explained by any individual 
haplotype (in this case CAST). For reference, the arrow below 
this panel runs from the TSS of \textit{Pkd2} (vertical bar) to the 
TES (arrow head) and shows the direction of transcription. The 
gray arrows at the top indicate two regions of interest where 
chromatin state explains height amounts of variance in gene expression. 
(B) The chromatin states assigned to each 200 bp window in this gene 
for each inbred mouse strain. States are colored by their association 
with gene expression in the inbred mice. Red indicates a positive 
association with gene expression, and blue indicates a negative 
association. Each row shows the chromatin states for a single inbred 
strain, which is indicated by the label on the left. (C) SNPs along 
the gene body for each inbred strain. The reference genotype is shown 
in gray. SNPs are colored by genotype as shown in the legend. (D) Percent 
DNA methylation for each inbred strain along the \textit{Pkd2} gene body. 
Percentages are binned into 0\% (blue) 50\% (yellow) and 100\% (red). (E) 
Association of haplotype with expression of \textit{Pkd2} in the DO. Haplotype 
effects are colored by from which each allele was derived. (F) \textit{Pkd2} 
expression levels across inbred mouse strains. For ease of comparison, 
all panels B through F are shown in the same order as the haplotype effects.}
\label{fig:example_gene}
\end{figure}

Both chromatin state and SNPs in these two regions were strongly 
associated DO expression levels of \textit{Pkd2} (Fig. \ref{fig:example_gene}A). 
Comparing these regions marked in panel A to the chromatin states
in panel B, we see that these two regions both have activating 
states in the high-expressing haplotypes and an absence of activating
marks in the low-expressing haplotypes. We therefore hypothesized that 
these two regions are heritable cis-regulatory regions for \textit{Pkd2}.

The spatial patterns in the SNPs partially mirror those in chromatin 
state (Fig. \ref{fig:example_gene}C). SNPs underlying the more proximal 
enhancer region could potentially influence gene expression by altering 
local chromatin state. However, the more distal putative cis-regulatory
region has no underlying SNPs, suggesting that there is an alternative 
mechanism for determining chromatin state at this location. Perhaps SNPs 
in the TSS region regulate chromatin state in both regions. For this 
particular gene, variation in DNA methylation was not associated with 
\textit{Pkd2} expression in the DO (Fig. \ref{fig:example_gene}D). 


# Discussion

In this study we showed that the epigenetic landscape
of hepatocytes varied widely across commonly used inbred
mouse strains and that this variation was associated with 
strain differences in gene expression. We saw evidence that
both chromatin state defined by combinatorial histone modifications,
as well as DNA methylation, were heritable mechanisms contributing 
to inter-individual variation in gene expression in mice. For DNA 
methylation, heritable variation was drivin in part by strain-specific 
CpG sites. These CpG sites were enriched in enhancers, specifically, 
weak, strong, and poised enhancers distal to the TSS. Strain-specific 
CpG sites were depleted in promoter regions and CpG islands suggesting 
that these regions are more highly conserved across the inbred strains 
studied here and that enhancer regions are the most diverged. This 
divergence of CpG sites in enhancer regions results in small varation 
in gene expression across strains relative to potentially large 
or catastrophic changes that might be expected with loss or gain of 
CpG sites in promoter regions.

The chromatin states we identified were represented by combinations of 
histone modifications that were enriched around previously predicted 
chromatin states in mouse liver. We used these enrichments to annotate 
each state, but noted that the annotations agreed both with relative 
abundance around the gene body and with associations to gene expression. 

Five of the `r num.states` state we identified were enriched around
bivalent promoters. Bivalent states are characterized by a 
combination of activating and repressing histone modifications
\citep{pmid23788621, pmid22513113}. Consistent with this definition,
all five states included the repressive mark, H3K27me3, and at least 
one of the activating marks. All of these states were also most 
abundant around the gene TSS, further supporting the annotation of 
promoter. Three of these states, States 10, 11, and 12, were associated 
with reduced gene expression both across genes and across strains, 
suggesting that these states marked genes that were poised for expression, 
but were not highly expressed. These associations were replicated in the DO 
for States 11 and 12, suggesting that these states represented a heritable 
form of gene expression regulation.

Bivalent promoters are typically considered dynamic states that change 
over the course of differentiation and in response to external stimuli. 
These regulatory regions have been studied primarily in the context 
of development. They are abundant in undifferentiated cells, and are 
often resolved either to active promoters or to silenced promoters as 
the cells differentiate into their final state \citep{pmid23788621, 
pmid22513113}. These promoters have also been shown to be important 
in the response to changes in the environment--their abundance increases 
in breast cancer cells in response to hypoxia \citep{pmid27800026}. It 
is therefore interesting to see apparently heritable bivalent promoters 
in differentiated hepatocytes. Genes marked by State 11 were enriched for 
mesodermal cell differentiation and Notch signaling suggesting a developmental 
role for this state. Similarly, genes marked by State 12 were enriched 
for blood vessel and endothelial morphogenesis as well as Wnt signaling. 

That we identified these states in differentiated hepatocytes may 
indicate that a subset of developmental genes retain the ability 
to be activated under certain circumstances, such as during liver 
regeneration in response to injury. Both Wnt signaling and Notch
signaling are involved in wound repair \citep{shi2015wnt, 
chigurupati2007involvement, whyte2012wnt} and liver regeneration
\citep{yue2018role, hu2021wnt, thompson2007wnt}. The observation
that these states likely represent a heritable form of cis-regulation 
is intriguing and may suggest heritable variation in response to 
liver injury or convergent evolution of regeneration pathways.

State 5, which was also annotated as a bivalent promoter, was 
interesting in that it was associated with lower gene expression
in hepatocytes, but with higher gene expression when looking 
across strains. That is, genes with State 5 were more lowly 
expressed than other hepatocyte genes, but for any given gene, 
strains with State 5 had higher expression than other strains.

The association of State 5 with reduced expression within hepatocytes 
is consistent with the annotation of bivalent promoter. Genes marked 
with this state were enriched for vascular development and Wnt signaling, 
further supporting the annotation. When positions marked with State 5 
varied across strains, the most common alternate state at these positions 
was State 12, another bivalent promoter. Thus, this group of genes in 
general was regulated by bivalent promoters and down-regulated relative 
to other genes. However, our results suggest that State 5 was associated 
with less severe downregulation when compared with State 12, resulting 
in an apparent upregulation when looking across strains. It is also 
possible that State 5 is a mixture of State 12 and other states and 
therefore produced inconsistent results. State 5 had a very similar 
abundance distribution, effect size distribution, and GO term enrichments 
to State 12. As a whole, the group of states annotated as bivalent promoters 
raise the intriguing possibility of identifying new modes of expression 
regulation through histone modification. Although these five states all 
recieved the same annotation, each had a unique pattern of distribution 
around the gene body and association with gene expression suggesting that 
each represents a different functional element in the mouse genome.

The diversity in the associations with gene expression observed across 
all `r num.states` chromatin states highlights the importance of analyzing 
combinatorial states as opposed to individual histone modifications. The 
three states with the largest positive associations with transcription each 
had a distinct combination of the three activating histone marks: H3K4me1, 
H3K4me3, and H3K27ac. And although all three states were associated with 
increased gene expression, each had a distinct spatial distribution. This 
variation in spatial distribution was mirrored in the spatial associations 
with transcription. The distinct patterns among these states would not be 
detectable without analysis of the histone modifications in combination. 
These results highlight the complexity of the histone code and the 
importance of analyzing combinatorial states.

State `r unknown.state` further illustrates the importance of the 
combinatorial approach. State `r unknown.state` was defined as the 
presence of H3K4me3 and the absence of all other marks. H3K4me3 is 
most frequently associated with increased transcriptional activity 
\citep{pmid15680324, pmid14661024, pmid12353038, pmid16728976}, so the 
association of state `r unknown.state` with reduced transcription is a 
deviation from the dominant paradigm. This state was enriched around
predicted poised enhancers in mouse liver data, and genes marked with 
this state were enriched for functions such as stress response, DNA 
damage repair, and ncRNA processing. Taken together, these results
suggest that this state may be used to regulate subsets of genes 
involved in responses to environmental stimuli. They further demonstrate
that the relationship between H3K4me3 and gene expression is more 
complex that simple activation. 

The merging of DO expression quantitative trait loci with inbred
chromatin state maps offers an interesting way to identify 
cis-regulatory regions. The \textit{Pkd2} example illustrates 
how this could be done. Given that there is a cis-eQTL at this 
locus, and that imputed chromatin state explained a large amount 
of variance in DO gene expression, we looked more closely at chromatin 
state within the inbred founders to identify putative cis-regulatory 
regions (Fig. \ref{fig:example_gene}A arrows). The variation in chromatin 
state across the inbred strains in these regions is highly correlated 
with gene expression in the outbred animals, pointing to possible 
molecular mechanisms for the observed eQTL. The PWK and WSB haplotypes 
lack the activating states seen in the other six haplotypes. This
lack of activating states was correlated with reduced gene expression 
in these strains. Further, the haplotypes with broader activating 
states also appeared to have higher \textit{Pkd2} expression 
suggesting that the width of the activating state may also contribute
to increased gene expression. The CpG sites in and arround these putative
regulatory regions are unmethylated across all strains, further supporting 
the hypothesis chraomtin state in these regions is actively regulating 
transcription. Validation of these regions is beyond the scope of this 
study, but our results suggest that combining DO eQTL data with inbred 
epigenetic data may serve as an important resource in identifying putative 
regulatory regions.

The discordance between the patterns of chromatin state and 
SNPs in this gene is particularly interesting. Variation in 
chromatin state at the intragenic enhancer is present in the 
absence of local SNPs. This suggests that the presence of the 
downstream enhancer is determined by another mechanism, perhaps 
SNPs acting in *trans* to this region, or local variation that 
was not measured by SNP genotyping, e.g. indels. Genetic 
variation located at a distance from the putative enhancer sites
could also potentially alter the 3D configuration of the genome
resulting in variable access of transcription factors to the 
enhancer. 

Broadly, local variation in chromatin state, DNA methylation, and 
individual SNPs, were all more highly correlated with DO gene 
expression than individual haplotypes were. Individual haplotypes
are a measure of ancestry, whereas chromatin state, DNA methylation,
and SNPs all potentially functionally related to gene expression. 
Two haplotypes that are not identical by descent may share a 
repressor state that is functionally associated with reduced gene 
expression. These observations raise the possibility
of shifting toward mapping traits with functional elements of the 
genome rather than ancestral allele labels. Many researchers already
use SNPs in mapping rather than haplotype, but the set of functional
features could be expanded further to include DNA methylation and 
histone modifications. By combining the power of haplotype mapping 
with the high resolution and mechanistic insights of other genomic
and epigenomic features, we can begin to build mechanistic hypotheses 
that link genetic variation to variation in gene expression and physiology. 

# Materials and Methods

## Ethics Statement

All animal procedures followed Association for Assessment and Accreditation of 
Laboratory Animal Care guidelines and were approved by Institutional Animal 
Care and Use Committee (The Jackson Laboratory, Protocol AUS #04008).

## Inbred Mice

Three female mice from each of nine inbred strains were used.
Eight of these strains (129S1/SvImJ, A/J, C57BL/6J, CAST/EiJ,
NOD/ShiLtJ, NZO/HlLtJ, PWK/PhJ, and WSB/EiJ) are the eight 
strains that served as founders of the Collaborative 
Cross/Diversity Outbred mice \citep{Chesler:2008ge}. The ninth strain, DBA/2J, 
will facilitate the interpretation of existing and forthcoming 
genetic mapping data obtained from the BxD recombinant inbred 
strain panel. Samples were harvested from the mice at 12 
weeks of age.

### Liver perfusion

To purify hepatocytes from the liver cell population, the mouse livers 
were perfused with 87 CDU/mL Liberase collagenase with 0.02\% CaCl2 in 
Leffert's buffer to digest the liver into a single-cell suspension, 
and then isolated using centrifugation. 

We aliquoted $5 \times 10^{6}$ cells for each RNA-Seq and bisulfite sequencing, 
and the rest were cross-linked for ChIP assays. Both aliquots were spun down
at 200 rpm for 5 min, and resuspended in $1200\mu L$ RTL+BME (for RNA-Seq) 
or frozen as a cell pellet in liquid nitrogen (for bisulfite sequencing). 
In the sample for ChIP-Seq, protein complexes were cross-linked to DNA
using 37\% formaldehyde in methanol. All cell samples were stored at -80C 
until used (See Supplemental Methods for more detail). 

### Hepatocyte histone binding and gene expression assays
 
Hepatocyte samples were used in the following assays:
 
1. RNA-seq to quantify mRNA and long non-coding RNA expression, with 
  approximately 30 million reads per sample. 
2. Reduced-representation bisulfate sequencing to identify methylation 
  states of approximately two million CpG sites in the genome. The average 
  read depth was 20-30x. 
4. Chromatin immunoprecipitation and sequencing to assess binding of the 
  following histone marks: 
    a. H3K4me3 to map active promoters 
    b. H3K4me1 to identify active and poised enhancers 
    c. H3K27me3 to identify polycomb repression
    d. H3K27ac, to identify actively used enhancers 
    e. A negative control (input chromatin) 

Samples were sequenced with $\sim40$ million reads per sample.

The samples for RNA-Seq in RTL+BME buffer were sent to The Jackson Laboratory 
Gene Expression Service for RNA extraction and library synthesis.

### Histone chromatin immunoprecipitation assays

After extraction, hepatocyte cells were lysed to release the nuclei,
spun down, and resuspended in 130ul MNase buffer with 1mM PMSF (Sigma, #78830)
and 1x protease inhibitor cocktail (Roche) to prevent histone protein degradation. 
The samples were then digested with 15U of micrococcal nuclease (MNase), 
which digests the exposed DNA, but leaves the nucleosome-bound DNA intact.
We confirmed digestion of nucleosomes into 150bp fragments with 
agarose gel. The digestion reaction was stopped with EDTA and samples 
were used immediately in the ChIP assay. The ChIP assay was performed with
Dynabead Protein G beads and histone antibodies (H3K4me3: Millipore #07-473,
H3K4me1: Millipore #07-436, H3K27me3: Millipore #07-449, H4K27ac: abcam ab4729). 
After binding to antibodies, samples were washed to remove unbound chromatin 
and then eluted with high-salt buffer and proteinase K to digest protein away 
from DNA-protein complexes. The DNA was purified using the Qiagen PCR 
purification kit. Quantification was performed using the Qubit quantification 
system (See Supplemental Methods).

## Diversity Outbred mice
We used previously published data from a population of 478 diversity outbred 
(DO) mice \citep{Svenson:2012hq}. DO mice (JAX:DO) are available from The 
Jackson Laboratory (Bar Harbor, ME) (stock number 009376).
The DO population included males and females from DO generations four 
through 11. Mice were randomly assigned to either a chow diet (6% fat by 
weight, LabDiet 5K52, LabDiet, Scott Distributing, Hudson, NH), or a high-fat, 
high-sucrose (HF/HS) diet (45\% fat, 40\% carbohydrates, and 15\% protein) 
(Envigo Teklad TD.08811, Envigo, Madison, WI). Mice were maintained on this 
diet for 26 weeks. 

To get the best estimates of the associations between epigenetic 
states and gene expression, we used all animals in the DO population
and regressed out the effects of sex and diet from all variables
before testing for associations. However, because the inbred animals 
used in this study were females maintained on a chow diet, it is possible
that either sex or diet could affect the results. To test whether sex
or diet had any effect on the associations between epigenetic features
and gene expression, we performed all tests using only females, and again 
only with chow-fed animals. Neither subsetting of the DO animals affected 
the results, so all results presented here use all animals.

### Genotyping
All DO mice were genotyped as described in Svenson \textit{et al.} (2012)
\citep{Svenson:2012hq} using the Mouse Universal Genotyping Array 
(MUGA) (7854 markers), and the MegaMUGA (77,642 markers) 
(GeneSeek, Lincoln, NE). All animal procedures were approved by the 
Animal Care and Use Committee at The Jackson Laboratory (Animal Use 
Summary \# 06006).

Founder haplotypes were inferred from SNPs using a Hidden Markov Model 
as described in \citet{Gatti:2014ko}. The MUGA and MegaMUGA arrays were
merged to create a final set of evenly spaced 64,000 interpolated markers. 

### Tissue collection and gene expression
At euthenasia, whole livers were collected and gene expression was
measured using RNA-Seq as described perviously \citep{pmid27309819, pmid28592500}. 
Briefly, hepatocyte RNA was isolated using the Trizol Plus 
RNA extraction kit (Life Technologies), and 100-bp single-end reads 
were generated on the Illumina HiSeq 2000.


## Data Processing

### Sequence processing

The raw sequencing data from both RNA-Seq and ChIP-Seq were put through 
the quality control program FastQC (0.11.5), and duplicate sequences 
were removed before downstream analysis. 

### Transcript quantification
Transcript sequences were aligned to strain-specific 
pseudo-genomes \citep{pmid27309819}, which integrate SNPs and 
indels from each strain based on the GRCm38 mouse genome 
build. The B6 samples were aligned directly to the reference 
mouse genome. The pseudogenomes were created using g2gtools 
(\url{http://churchill-lab.github.io/g2gtools/#overview}). 
We used EMASE (\url{https://github.com/churchill-lab/emase}) 
\citep{pmid29444201} to quantify the gene expression counts 
and DESeq2 vst transformation \citep{love2014moderated} to 
normalize the gene expression data. We filtered out transcripts 
with less than 1 CPM in two or more replicates.

### ChIP-Seq quantification
We used MACS 1.4.2 \citep{pmid18798982} to identify peaks in the ChIP-Seq 
sequencing data, with a significance threshold of $p \leq 10^{-5}$. 
In order to compare peaks across strains, we converted the MACS output 
peak coordinates to common B6 coordinates using g2g tools. 

## Quantifying DNA methylation
RRBS data were processed using a bismark-based pipeline modified from 
\citep{pmid30348905}. The pipeline uses Trim Galore! 0.6.3 
\url{https://www.bioinformatics.babraham.ac.uk/projects/trim_galore/} for QC, 
followed by the trimRRBSdiversityAdaptCustomers.py script from NuGen for 
trimming the diversity adapters. This script is available at: 
\url{https://github.com/nugentechnologies/NuMetRRBS}

All samples had comparable quality levels and no outstanding flags. 
Total number of reads was 45-90 million, with an average read length 
of about 50 bp. Quality scores were mostly above 30 (including error bars), 
with the average above 38. Duplication level was reduced to $<2$ for about 
95\% of the sequences. 

High quality reads were aligned to a custom strain pseudogenomes, using 
bowtie2 as implemented in Bismark 0.22 \citep{pmid21493656}. The pseudogenomes were
created by incorporating strain-specific SNPs and indels into the reference 
genome using g2gtools, allowing a more precise characterization of methylation 
patterns. Bismark methylation extractor tool was then used for creating a bed 
file of estimated methylation proportions for each animal, which was then translated 
to the reference mouse genome (GRCm38) coordinates using g2gtools. Unlike other 
liftover tools, g2gtools does not throw away alignments that land on indel regions. 
B6 samples were aligned directly to the reference mouse genome (mm10).

## Analysis of histone modifications

### Identification of chromatin states
We used ChromHMM (1.22) \citep{pmid29120462} to identify 
chromatin states, which are unique combinations of the four 
chromatin modifications, for example, one state could consist 
of high levels of both H3K4me3 and H3K4me1, and low levels 
of the other two modifications. We conducted all subsequent 
analyses at the level of the chromatin state. 

Prior to running ChromHMM, we converted the bam files that had
been aligned to the B6 genome as described above to bed files
using the bedtools function bamtobed \citep{quinlan2010bedtools}.
We then binarized the bed files using the BinarizeBed function
in ChromHMM with default parameters. 

We calculated chromatin states for all numbers of states between 
four and 16, which is the maximum number of states possible with four 
binary chromatin modifications ($2^n$). We ran all mouse strains
together in the same model as if they were different cell types
in a standard run of ChromHMM.

To ensure we were analyzing the most biologically meaningful chromatin 
states, we aligned states across all models of four to 16 states 
by assigning each to one of the sixteen possible binary states using an 
emissions probability of 0.3 as the threshold for presence/absence of the
histone mark. This threshold was used for comparison purposes only, and 
produced the most stable state estimates between models. We then 
investigated the stability of three features across all states: the 
emissions probabilities (Supplemental Fig. \ref{supp_fig:model_emissions_comparison}), 
the abundance of each state across transcribed genes (Supplemental Figure 
\ref{supp_fig:model_abundance_comparison}), and the associations of each state 
with transcription (Supplemental Fig. \ref{supp_fig:model_effect_comparison}). 
Methods for each of these analyses are described separately below. All 
measures were remarkably consistent across all models, but the 
`r num.states`-state model was characterized by a wide range of 
relatively abundant states with relatively strong associations with 
expression. We used this model for all subsequent analyses. For more 
details on how the different models were compared, see Supplemental Methods.

### Genome distribution of chromatin states

We investigated genomic distributions of chromatin states using the 
ChromHMM function OverlapEnrichment to calculate enrichment of each 
state around known functional elements in the mouse genome. We 
analyzed the following features:

* Predicted Liver Chromatin States - We downloaded predicted liver 
chromatin states through the UCSC genome browser on February 14, 2023 
(http://genome.ucsc.edu/cgi-bin/hgTables). We selected Expression 
and Regulation -> Chromatin State -> cHMM liver P0 (encode3RenChromHmmLiverP0) 
under the mouse mm10 assembly. These data include chromatin state 
annotations for mouse liver on post-natal day 0. The annotations 
were based on ChIP-Seq measurements of eight histone modifications: 
H3K27ac, H3K27me3, H3K4me3, H3K4me2, H3K4me1, H3K9me3, H3K9ac, and 
H3K36me3. ChromHMM was used to identify 15 chromatin states that were 
each annotated with a putative function based in the literature.

* CpG Islands - Annotations of CpG islands in the mouse genome were included 
with the release of ChromHMM.

* Intergenic - Annotations of intergenic regions in the mouse genome were
included with the release of ChromHMM.

## Gene body distribution of chromatin states

In addition to these enrichments around individual elements, we also calculated 
chromatin state abundance relative to the main anatomical features of a gene.
For each transcribed gene, we normalized the base pair positions to the length
of the gene such that the transcription start site (TSS) was fixed at 0, and the
transcription end site (TES) was fixed at 1 taking into account the encoding strand
of DNA. We also included 1000 bp upstream of the TSS and 1000 bp downstream of the 
TES, which were converted to values below 0 and above 1 respectively. To map chromatin 
states to the normalized positions, we binned the normalized positions into 42 bins 
running from -0.5 to 1.5. This range included some upstream and downstream regions 
around the gene body and gave us good resolution around 0 and 1. If a bin encompassed 
multiple positions in the gene, we assigned the mean value of the feature of interest 
to the bin. To avoid potential contamination from regulatory regions of nearby genes, 
we only included genes that were at least 2kb from their nearest neighbor, for a final 
set of 14,048 genes.

### Chromatin state and gene expression

We calculated the association of each chromatin state with gene expression 
(Fig. \ref{fig:state_overview}C). We did this both across genes
and across strains. The across-gene analysis identified states that 
are associated with high expression and low expression within the 
hepatocytes. The across-strain analysis investigated whether 
variation in chromatin state across strains was associated with 
variation in gene expression across strains. 

For each transcribed gene, we calculated the proportion of the gene 
body that was assigned to each chromatin state. We then fit a linear 
model separately for each state to calculate the association of state 
proportion with gene expression:

\begin{equation}\label{eqn:model}
y_{e} = \beta x_{s} + \epsilon
\end{equation}

where $y_{e}$ is the rank normal scores \citep{conover1999practical} of the full 
transcriptome in a single inbred strain, and $x_{s}$ is the rank normal
proportion of each gene that was assigned to state $s$. We fit this model 
for each strain and each state to yield one $\beta$ coefficient with a 95\% 
confidence interval. We fit the strains independently to better identify
variation in chromatin state effects across strains. However, the effects 
were not different across strains (ANOVA $p > 0.5$), so we averaged the 
effects and confidence intervals across strains to yield one summary effect 
for each state. We further fit models for each state independently, rather
than using multiple regression, because we were primarily interested in the
marginal effects of each state for this study.

To calculate the association of each chromatin state with gene expression 
across strains, we first standardized transcript abundance across strains 
for each transcript. We also standardized the proportion of each chromatin 
state for each gene across strains. We then fit the same linear model, 
where $y_{e}$ was a rank normal vector concatenating all standardized 
expression levels across all strains, and $x_{s}$ was a rank normal vector 
concatenating all standardized state proportions across all strains. We fit 
the model for each state independently yielding a $\beta$ coefficient and 
95\% confidence interval for each state. 

In addition to calculating the association of state proportion across
the full gene body with gene expression, we also performed the same 
calculations in a position-based manner (Fig. \ref{fig:state_effects}). 
To do this, we normalized the genomic positions of all chromatin 
states to run between 0 at the transcription start site (TSS) and 1 at 
the transcription end site (TES) as described above. In dividing chromatin 
state values into bins, we averaged all positions for each state that were 
contained in each bin. We fit the linear model described above for each 
positional bin thus creating position-based effect sizes for chromatin 
state on gene expression across genes and across strains.


## Analysis of DNA methylation

### Creation of DNA methylome
We combined the DNA methylation data into a single methylome cataloging
all unique methylated sites across all strains. For each site, we averaged 
the percent methylation across the three replicates in each strain. The final
methylome contained 5,311,670 unique CpG sites across the genomes of all nine 
strains. Because methylated CpG sites can be fully methylated, unmethylated, 
or hemi-methylated, we rounded the average percent methylation at each site 
to the nearest 0, 50, or 100\%.

### Decomposition of DNA methylome
To calculate the DNA methylation similarity across individuals 
shown in Figure \ref{fig:pc_plots}B we used the subset of the 
CpG sites that were shared across all strains at each B6 reference 
position. The resulting matrix contained individual mice in columns 
and shared methylation sites in rows. Each cell contained the measured 
level of DNA methylation at that position. We performed principal 
components analysis on this matrix.

### Strain-specific CpG sites
In addition to the analysis of CpG sites that were shared across
genes, we analyzed CpG sites that were strain-specific. We defined
a strain-specific CpG site as one that was present in all members
of at least one strain and absent in all members of at least one
other strain. 

### Distribution and methylation of CpG sites
We used the enrichment function in ChromHMM described above to identify 
enrichment of CpG sites around functional elements (e.g. CpG islands, 
mouse liver enhancers, and mouse liver promoters). These features 
are described above in the section "Genome distribution of 
chromatin states." We further performed position-based analyses
of both CpG density and percent methylation similar to the 
position-based abundance analyses performed for chromatin states.

To calculate overall CpG density relative to gene bodies, we 
calculated the inverse of the inter-CpG base pair distances within 
1kb of each expressed gene. We then normalized the position of 
each CpG to reflect its position relative to the gene's TSS 
(at 0) and its TES (at 1) as described above. We took the average 
of these values in each of 42 bins running from a relative position 
of -0.5 to 1.5 Figure \ref{fig:state_abundance}C shows the average 
inverse inter-CpG distance across all 42 bins. CpG sites were 
most densely packed near the TSS (relative gene position = 0)
as expected. 

Figure \ref{fig:state_abundance}D shows the average percent 
methylation in each of these bins, which was calculated in the 
same manner as above but we calculated the median percent methylation
in each bin rather than the inverse inter-CpG distance. The figure 
shows that CpG sites tended to be unmethylated near the TSS as expected.

### Association of DNA methylation with gene expression
As with chromatin state, we assessed the association between DNA 
methylation and gene expression both across genes 
(Fig. \ref{fig:DNA_methylation_effect}A) and across strains 
(Fig. \ref{fig:DNA_methylation_effect}B). As with chromatin state,
we binned the normalized CpG positions into 42 bins running from just
upstream of the TSS to just downstream of the TES. We treated missing
CpG sites in individual strains as unmethylated, as it is uncommon
for non-CpG sites to be methylated. This allowed us to test strain-specific
CpG sites and variation in DNA methylation percent simultaneously. We 
then fit the linear model shown in equation 1 where $x_{s}$ was the 
rank normal percent methylation either across genes or across 
strains in each position bin. Because the effect of DNA methylation 
on gene expression is well-known to be dependent on position, we only 
calculated a position-dependent association with expression. We did not 
calculate the association of percent methylation across the full gene 
with expression.

### Interactions between chromatin state and DNA methylation
We repeated the above analyses for DNA methylation conditioned
on each of the `r num.states` chromatin states. To do this, we 
isolated all CpG sites that were contained in the genomic regions 
defined by each chromatin state. We then performed the above analysis 
on each subset of CpG sites independently.

## Imputation of genomic features in Diversity Outbred mice
To assess the extent to which chromatin state and DNA methylation 
were associated with local expression QTLs, we imputed local 
chromatin state and DNA methylation into the population of 
diversity outbred (DO) mice. We compared the effects of the imputed 
epigenetic features to imputed SNPs and to local haplotype effects
as measured in the DO.

All imputations followed the same basic procedure: For each transcript, 
we identified the haplotype probabilities in the DO mice at the genetic 
marker nearest the gene transcription start site. This matrix held DO 
individuals in rows and DO founder haplotypes in columns 
(Supplemental Fig. \ref{supp_fig:imputation}).

For each transcript, we also generated a three-dimensional array 
representing the genomic features (chromatin state, DNA methylation 
status, or SNP genotype) derived from the DO founders. This array 
held DO founders in rows, feature state in columns, and genomic 
position in the third dimension. The feature state for chromatin 
consisted of states one through `r num.states`, for SNPs feature state 
consisted of the genotypes A,C,G, and T.

We then multiplied the haplotype probabilities by each genomic feature 
array to obtain the imputed genomic feature for each DO mouse. This final 
array held DO individuals in rows, the genomic feature in the second 
dimension, and genomic position in the third dimension 
(Supplemental Fig. \ref{supp_fig:imputation}). This array is analagous 
to the genoprobs object in R/qtl2 \citep{pmid30591514}. The genomic 
position dimension included all positions from 1 kb upstream of the 
TSS to 1 kb downstream of the TES for the given transcript. SNP data 
for the DO founders in mm10 coordinates were downloaded from the Sanger 
SNP database \citep{keane2011mouse} on July 6, 2021.

To calculate the association between each imputed genomic feature and 
gene expression in the DO population, we fit a linear model 
$y_{e} = \beta x_{s} + \epsilon$ where $y_{e}$ was DO gene 
expression of a single transcript, and $x_{s}$ was the 
imputed level of a single chromatin state at a single base 
pair position within the encoding gene of the transcript. 
Prior to fitting this model, we regressed sex and DO generation 
out from all variables so that they would not be included in the
estimate of variance explained by each chromatin state.

Testing each state seprately is a bit artificial, since no single 
haplotype will explain as much variance as using all haplotypes 
together in a multiple regression. However, it was critical in 
this study to maintain a single degree of freedom across all 
features so that we could compare them. Otherwise haplotypes 
have seven degrees of freedom (df) at each location, chromatin 
states potentially have 13 df, although in practice they 
typically have between two and four df, and both SNPs and DNA 
methylation have only one df. Thus, to compare the features, we 
tested only a single state at a time. From these linear models, 
we calculated the variance explained ($R^2$) by each genomic 
feature at each position (Fig. \ref{fig:effect_distrubutions}), 
thereby relating gene expression in the DO to each position of 
the imputed feature in and around the gene body. We also kept 
the $\beta$ coefficients to identify overall trends in positive 
or negative associations on gene expression for each genomic 
feature at each position (Fig. \ref{fig:state_effects}C).

### Permutations

Because any feature imputed from haplotype will be correlated with 
any feature that haplotype is correlated with, we performed 
permutations of the above statistics to assess whether each genomic 
feature was significantly correlated with gene expression beyond the 
effect of the imputation itself. To do the permutations, we shuffled 
the strain labels on each genomic feature vector (chromatin states, 
DNA methylation percent, or SNPs). This randomized the association 
between haplotype and the assigned genomic feature while preserving 
the association between haplotype and gene expression. We then 
re-imputed the permuted features into the DO and performed the 
association tests on the randomized imputed values as described above. 

We performed 1000 permutations for each transcript retaining the
$R^2$ value from each permutation. We then calculated an empirical 
$p$-value for the $R^2$ of each transcript based on these permutations. 
This was the number of times the permutations met or exceeded the 
observed $R^2$ value divided by the total number of permutations.
We then analyzed the empirical $p$-value distributions for uniformity.
A uniform $p$-value distribution across the transcripts would suggest 
that the given genomic feature was not significantly associated with 
gene expression. An enrichment of small $p$-values, on the other hand, 
would suggest that there is a significant association between the imputed 
genomic feature and gene expression beyond that conferred by the imputation 
itself. The $p$-value distributions for all three genomic features were 
highly enriched for small $p$-values (all Kruskal-Wallis $p < 2^{-16}$), 
suggesting that, although many individual imputed values were not significantly 
associated with gene expression, overall each genomic feature could 
be significantly associated with gene expression (Supplemental Fig. 
\ref{supp_fig:empirical_p}).

# Data Access
All raw and processed sequencing data generated in this study have 
been submitted to the NCBI Gene Expression Omnibus 
(GEO; https://www.ncbi.nlm.nih.gov/geo/) under accession 
number GSE213968.

Code to run the analyses in this study are at 
\url{https://github.com/annaLtyler/Epigenetics_Manuscript}.

# Competing Interest Statement
The authors do not have any competing interests to declare.

# Acknowledgements
This work was funded by The Jackson Laboratory Director's 
Innovation Fund and the National Institutes of Health grants 
R01 GM115518 (to G.W.C), GM070683 (to G.A.C), R35 GM133724 
(to C.L.B.), and P30 CA034196. J.J.T. is a Scholar of the 
Leukemia & Lymphoma Society.
 
We gratefully acknowledge expert assistance from Genome 
Technologies, Gene Expression Services, and Information 
Technologies at The Jackson Laboratory.

\pagebreak

# Supplemental Figure Legends

\beginsupplement

\begin{figure}[ht!]
\caption{Correlations of multiple genomic features across inbred
mice. Each panel shows one genomic feature A. Transcriptome;
B. Methylome; C. H3K27ac; D. H3K27me3; E. H3K4me1; F. H3K4me3.
These figures complement Figure 1 in the manuscript and show
a more finely detailed structure of the correlations between
individuals and strains.
}
\label{supp_fig:feature_correlation}
\end{figure}

\begin{figure}[ht!]
\caption{These figures are companions to those shown in Figure 
\ref{fig:state_abundance}A. Each panel shows the abundance of 
a single state for each strain. The color of each line indicates 
the strain. The abundance pattern of each state is largely identical 
across the strains. Deviations are seen only outside the gene body
for very rare states are are thus artefacts due to low numbers
of example genes.
}
\label{supp_fig:state_abundance_by_strain}
\end{figure}


\begin{figure}[ht!]
\caption{This figure is a companion to the DNA methylation 
percent panel in Figure \ref{fig:state_abundance}D. It shows 
DNA methylation percent for each strain along the gene body. 
Strain is indicated by color.
}
\label{supp_fig:percent_methyl_by_strain}
\end{figure}

\begin{figure}[ht!]
\caption{The following figure shows the association of each chromatin 
state to gene expression for reference in the first colum. The
second column shows the percent DNA methylation of CpG sites in 
each state. The third column shows the association between DNA 
methylation in each state with gene expression. The fourth column
shows the state annotation for reference. Note that DNA methylation
in each state tended to be associated with expression in a manner
opposite to that of the chromatin state. If the chromatin state
had a positive association with gene expression, DNA methylation 
in that state had a negative association with gene expression and
vice versa.
}
\label{supp_fig:methyl_effect_by_state}
\end{figure}


\begin{figure}[ht!]
\caption{Comparison of emissions probabilities across all ChromHMM models.
Each row contains data for a single ChromHMM model fit to the number of
states indicated on either side of the row. Each set of four columns shows
data for each of the four histone modifications. Each set is separated from
the next by a column of gray for ease of visualization. The bottom row, the
reference row, shows the ideal state that all model states are being compared
to. Blue indicates absence of the histone mark and red indicates presence.
For each ChromHMM model, each state was assigned to one of the reference states
using an emissions probability of 0.3 as a threshold for presence of the histone
modification. If a state was not present in the given model, the corresponding
area is shown in gray. Emissions probabilities near 0 are shown in blue, and 
probabilities near 1 are shown in red. Orange and yellow indicate intermediate
probabilities. Aligning the states across all models shows a remarkable 
stability in the emissions across models, seen as vertical bars of consistent 
color.}
\label{supp_fig:model_emissions_comparison}
\end{figure}


\begin{figure}[ht!]
\caption{Comparison of state abundance across all ChromHMM models. The left-most column
shows the annotation for each state. Unannotated states are marked with a dash. The 
binary heatmap indicates which histone modifications were present in each state: 1 
indicates presence, and 0 indicates absence. The histone modifications are labeled
at the bottom of each column. The continuous heatmap shows the abundance of each state
(in rows) in each ChromHMM model (in columns). The abundance is the proportion of 
transcribed genes with the state present. Less abundant states are shaded blue, and 
more abundant states are shaded yellow, orange, and red. The number of states in the
model is indicated at the bottom of each column. The black box highlights the model
used in this study -- the `r num.states`-state model. State abundance was remarkably
stable across the different models.}
\label{supp_fig:model_abundance_comparison}
\end{figure}

\begin{figure}[ht!]
\caption{Comparison of state associations with gene expression across all 
ChromHMM models. This figure is identical to Figure 
\ref{supp_fig:model_abundance_comparison}, except that the cells in the 
continuous heatmap show the effect size of each state on gene expression
across all ChromHMM models. The effect was the $\beta$ coefficient derived from
a linear model. Similar to state abundance, the associations were remarkably 
stable across models.}
\label{supp_fig:model_effect_comparison}
\end{figure}

\begin{figure}[ht!]
\caption{Schematic for imputation of histone modifications into the 
DO mice. For a single transcript imputation was made by multiplying 
a three-dimensional array, containing chromatin state by strain by
position, by a two-dimensional array, contatining haplotype probabilities
by DO individual, to create a three-dimensional array, containing 
individual by position by chromatin state probability.}
\label{supp_fig:imputation}
\end{figure}


\begin{figure}[ht!]
\caption{This figure shows empirical $p$-value distributions obtained
by permuting imputed genomic features and associating them
with gene expression in the DO mice. All distributions are
enriched with small values suggesting that overall each 
genomic feature   is associated with gene expression in the DO
beyond the simple correlation with haplotype.
}
\label{supp_fig:empirical_p}
\end{figure}


\pagebreak

\bibliography{epigenetics.bib}
